---
title: "Filter doublets and add sample information to C133_Neeland data"
subtitle: "Batches 0-6"
author: "Jovana Maksimovic"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
site: workflowr::wflow_site
output:
  bookdown::html_document2:
    base_format: workflowr::wflow_html
    toc: yes
    toc_float: yes
    theme: cosmo
    highlight: textmate
    number_sections: true
---

```{r, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

# Load libraries

```{r, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(BiocStyle)
  library(tidyverse)
  library(here)
  library(glue)
  library(patchwork)
  library(paletteer)
  library(readxl)
  library(ggupset)
  library(stringr)
})

```


# Load Data

```{r}
files <- list.files(here("data",
                         paste0("C133_Neeland_batch", 0:6),
                         "data", 
                         "SCEs"), 
                      pattern = "doublets_called",
                      full.names = TRUE)
  
sceLst <- sapply(files, function(fn){
    readRDS(file = fn)
})

sceLst
```


# Load sample information

```{r}
ext_metadata_df <- read_excel(
  here("data/CZI_samples_design_with_micro.xlsx"),
  col_types = 
    c("text", "numeric", "text", "text", "text", "numeric", "numeric", "text",
      "numeric", "text", "numeric", "numeric", "numeric", "numeric",
      "text", "text", "text")) %>%
  mutate(HASHTAG = paste0("Human_HTO_", HASHTAG)) %>%
  mutate(sample.id = ave(Donor_ID, Donor_ID, 
                         FUN = function(i) paste0("sample_", i, '.', seq_along(i)))) %>%
  dplyr::rename(Batch = BATCH, 
                Donor = Donor_ID, 
                Hashtag = HASHTAG,
                arest.id = AREST_ID) %>%
  mutate(Donor = as.factor(Donor))
knitr::kable(ext_metadata_df)
```


# Assign sample IDs to cells

For genetically distinct samples, cells are assigned to a sample using the genetic assignment; the sample information is then assigned using the corresponding HTO. For longitudinal samples *only*, both cells and sample information are assigned to the individual samples using HTOs. 

```{r}
sceLst <- sapply(1:length(sceLst), function(i){
  message(i)
  sce <- sceLst[[i]]
  
  # convert genetic and HTO labels to character
  sce$GeneticDonor <- as.character(sce$GeneticDonor)
  sce$dmmHTO <- ifelse(!stringr::str_detect(sce$dmmHTO, "Human_HTO_") & 
                         !(stringr::str_detect(sce$dmmHTO, "Doublet|Negative")),
                       paste0("Human_HTO_", sce$dmmHTO),
                       as.character(sce$dmmHTO))
  
  # match genetics and HTOs
  tab <- table(sce$GeneticDonor, sce$dmmHTO)
  tab <- tab[!rownames(tab) %in% c("Doublet","Unknown"),
             !colnames(tab) %in% c("Doublet","Negative")]
  # extract all genetically unique samples in this batch
  sub <- tab[rowSums(tab/rowSums(tab) > 0.2) == 1, , drop = FALSE]
  # create dictionary to translate genetics to HTO
  dict <- apply(sub, 1, function(x){
    colnames(tab)[which(x == max(x))]
  })
  dict <- c(dict, "Doublet" = "Doublet", "Unknown" = "Negative")
 
  sce$ProcessedHTO <- NA
  # assign HTO based on genetics
  genetics <- sce$GeneticDonor %in% names(dict)
  sce$ProcessedHTO[genetics] <- unname(dict[sce$GeneticDonor[genetics]])
  
  # if there are longitudinal samples also assign HTO based on hashtags
  if(dim(sub)[1] != dim(sub)[2]){
    HTO <- !sce$GeneticDonor %in% names(dict)
    sce$ProcessedHTO[HTO] <- sce$dmmHTO[HTO]
    
  }
  
  # add sample metadata based on processed HTO
  colData(sce) %>%
    data.frame() %>%
    rownames_to_column("cell") %>%
    left_join(ext_metadata_df %>% 
                dplyr::filter(Batch == (i-1)) %>%
                data.frame, 
              by = c("ProcessedHTO" = "Hashtag")) %>%
    mutate(sample.id = ifelse(is.na(sample.id), 
                              ProcessedHTO,
                              sample.id)) %>%
    column_to_rownames("cell") %>% 
    DataFrame() -> colData(sce)
  
  sce
})

```
## Number of cells per sample

```{r, fig.asp=1.75}
p <- lapply(1:length(sceLst), function(i){
  sce <- sceLst[[i]]
  colData(sce) %>%
    data.frame() %>%
    ggplot(aes(x = sample.id, fill = Donor)) +
    geom_bar(position = position_dodge()) + 
    geom_text(aes(label = ..count..),
              stat = "count", vjust = 1.5, 
              colour = "black", size = 2) +
    theme(legend.position = "none") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
    ggtitle(glue("batch {i-1}"))
})

wrap_plots(p, ncol = 2)
```

## No. of doublets called per batch

```{r, fig.asp=2}
p <- lapply(1:length(sceLst), function(i){
  sce <- sceLst[[i]]
  colData(sce) %>%
    data.frame() %>%
    mutate(scds = ifelse(hybrid_call, "Doublet", "Singlet"),
           scdf = ifelse(scDblFinder.class == "doublet", "Doublet", "Singlet")) %>%
    dplyr::select(ProcessedHTO, scds, scdf, Capture) %>% 
    rownames_to_column(var = "cell") %>%
    mutate(vireo_dmm_dbl = (ProcessedHTO == "Doublet"),
           scds_dbl = (scds == "Doublet"),
           scdf_dbl = (scdf == "Doublet")) %>%
    dplyr::select(cell, vireo_dmm_dbl, scds_dbl, scdf_dbl, Capture) %>% 
    pivot_longer(cols = c(vireo_dmm_dbl, scds_dbl, scdf_dbl), names_to = "method") %>%
    dplyr::filter(value == TRUE) %>%
    group_by(cell) %>%
    summarise(data = list(method), Capture) %>%
    rowwise() %>%
    ggplot(aes(x = data)) +
    geom_bar() +
    scale_x_upset(n_intersections = 20) +
    geom_text(stat = 'count', aes(label = after_stat(count)), 
              vjust = -0.5, size = 2) +
    ggtitle(glue("batch {i-1}"))
})

wrap_plots(p, ncol = 2)
```


# Doublet filtering

We HAVE to filter out cells that are called doublets by either `vireo` or `demuxmix` (for relevant samples only) because we have no singlet assignment to a sample for them. We also HAVE to filter out negative cells as they also have no singlet assignment. We will also filter out cells that are called doublets by BOTH `scds` and `scDblFinder` (both methods were suggested in the demuxafy paper) that are not already in the `vireo/demuxmix doublet pool`. The percentages for those doublets calls per capture are shown below.

```{r, results='asis'}
lapply(sceLst, function(sce){
  scds_scdf <- table(sce$Capture, sce$hybrid_call & sce$scDblFinder.class == "doublet" &
                       !sce$ProcessedHTO == "Doublet")
  vireo_dmm <- table(sce$Capture, sce$ProcessedHTO == "Doublet")
  
  if(length(levels(sce$Capture)) < 4) {
    data.frame(scds_scdf_dbl = scds_scdf[,2]/scds_scdf[,1]*100,
               vireo_dmm_dbl = vireo_dmm[,2]/vireo_dmm[,1]*100) -> tab
  } else {
    data.frame(scds_scdf_dbl = scds_scdf[,2]/scds_scdf[,1]*100) -> tab
  }
  
  tab %>% knitr::kable()
})
```

First, filter out *wheeze* and *CSLD* samples as we will focus only on *CF*.

```{r, fig.asp=1.75}
sceLst <- lapply(sceLst, function(sce){
  sce[, !sce$Disease %in% c("Wheeze", "CSLD")]
})

p <- lapply(1:length(sceLst), function(i){
  sce <- sceLst[[i]]
  colData(sce) %>%
    data.frame() %>%
    ggplot(aes(x = sample.id, fill = Donor)) +
    geom_bar(position = position_dodge()) + 
    geom_text(aes(label = ..count..),
              stat = "count", vjust = 1.5, 
              colour = "black", size = 2) +
    theme(legend.position = "none") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
    ggtitle(glue("batch {i-1}"))
})

wrap_plots(p, ncol = 2)
```
Then, filter out doublets.

```{r, fig.asp=1.75}
sceLst <- lapply(sceLst, function(sce){
  sce[, !(sce$ProcessedHTO %in% c("Doublet") | 
                                  (sce$hybrid_call & sce$scDblFinder.class == "doublet"))]
})

p <- lapply(1:length(sceLst), function(i){
  sce <- sceLst[[i]]
  colData(sce) %>%
    data.frame() %>%
    ggplot(aes(x = sample.id, fill = Donor)) +
    geom_bar(position = position_dodge()) + 
    geom_text(aes(label = ..count..),
              stat = "count", vjust = 1.5, 
              colour = "black", size = 2) +
    theme(legend.position = "none") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
    ggtitle(glue("batch {i-1}"))
})

wrap_plots(p, ncol = 2)
```


Finally, filter out negatives cells.

```{r, fig.asp=1.75}
sceLst <- lapply(sceLst, function(sce){
  sce[, !(sce$sample.id %in% "Negative")]
})

p <- lapply(1:length(sceLst), function(i){
  sce <- sceLst[[i]]
  colData(sce) %>%
    data.frame() %>%
    ggplot(aes(x = sample.id, fill = Donor)) +
    geom_bar(position = position_dodge()) + 
    geom_text(aes(label = ..count..),
              stat = "count", vjust = 1.5, 
              colour = "black", size = 2) +
    theme(legend.position = "none") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
    ggtitle(glue("batch {i-1}"))
})

wrap_plots(p, ncol = 2)
```

# Cell counts per sample after filtering

```{r, fig.asp=1.75}
p <- lapply(sceLst, function(sce){
  colData(sce) %>%
    data.frame() %>%
    mutate(donor.fct = factor(Donor, labels = 1:length(unique(Donor)))) %>%
    ggplot(aes(x = sample.id, fill = donor.fct)) +
    geom_bar(position = position_dodge()) + 
    geom_text(aes(label = ..count..),
              stat = "count", vjust = 1.5, 
              colour = "black", size = 2) +
    theme(legend.position = "none") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
})

wrap_plots(p, ncol = 2)
```

# QC per sample after filtering

```{r, fig.asp=3}
p <- lapply(sceLst, function(sce){
  colData(sce) %>%
    data.frame() %>% 
    mutate(donor.fct = factor(Donor, labels = 1:length(unique(Donor)))) %>%
    ggplot(aes(x = sample.id, y = sum,
               fill = donor.fct)) +
    geom_violin() +
    scale_y_log10() +
    theme(legend.position = "none") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
    labs(x = "Sample",
         y = "Library size (log10 scale)") -> p1
  
  colData(sce) %>%
    data.frame() %>% 
    mutate(donor.fct = factor(Donor, labels = 1:length(unique(Donor)))) %>%
    ggplot(aes(x = sample.id, y = detected,
               fill = donor.fct)) +
    geom_violin() +
    scale_y_log10() +
    theme(legend.position = "none") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
    labs(x = "Sample",
         y = "No. genes detected (log10 scale)") -> p2
  
  p1 + p2
})

wrap_plots(p, ncol = 1) &
  theme(axis.title.y = element_text(size = 9))
```


# Save data

```{r}
batches <- str_extract(files, "batch[0-6]")

sapply(1:length(sceLst), function(i){
  out <- here("data",
              paste0("C133_Neeland_", batches[i]),
              "data", 
              "SCEs", 
              glue("C133_Neeland_{batches[i]}.doublets_filtered.SCE.rds"))
  if(!file.exists(out)) saveRDS(sceLst[[i]], out)
  fs::file_chmod(out, "664")
  if(any(str_detect(fs::group_ids()$group_name, 
                    "oshlack_lab"))) fs::file_chown(out, 
                                                    group_id = "oshlack_lab")
})
```

# Session info {.appendix}














