---
title: "Integrate and cluster macrophage cells"
author: "Jovana Maksimovic and George Howitt"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
site: workflowr::wflow_site
output:
  bookdown::html_document2:
    base_format: workflowr::wflow_html
    toc: yes
    toc_float: yes
    theme: cosmo
    highlight: textmate
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries.

```{r}
suppressPackageStartupMessages({
 library(SingleCellExperiment)
 library(edgeR)
 library(tidyverse)
 library(ggplot2)
 library(Seurat)
 library(glmGamPoi)
 library(dittoSeq)
 library(here)
 library(clustree)
 library(patchwork)
 library(AnnotationDbi)
 library(org.Hs.eg.db)
 library(glue)
 library(speckle)
 library(tidyHeatmap)
 library(dsb)
})
```

# Load data

Load *macrophage* subset Seurat object.

```{r}
ambient <- ""
seu <- readRDS(here("data",
                    "C133_Neeland_merged",
                    glue("C133_Neeland_full_clean{ambient}_macrophages.SEU.rds")))
seu
```


# Data integration

Visualise batch effects.

```{r, message=FALSE, warning=FALSE, fig.asp=0.9}
seu <- ScaleData(seu) %>%
  FindVariableFeatures() %>%
  RunPCA(dims = 1:30, verbose = FALSE) %>%
  RunUMAP(dims = 1:30, verbose = FALSE)

DimPlot(seu, group.by = "Batch", reduction = "umap")
```

```{r, fig.asp=1}
DimPlot(seu, group.by = "predicted.ann_level_4", reduction = "umap") +
  theme(legend.position = "bottom",
        legend.direction = "vertical",
        legend.text = element_text(size = 10))
```

Examine cell library sizes after ambient removal per sample and per cell type. Some cells have very low library sizes after ambient removal.

```{r, fig.width=8}
VlnPlot(seu, features = "nCount_RNA", group.by = "sample.id", log = TRUE, pt.size = 0) + 
  NoLegend() + 
  geom_hline(yintercept = 250) +
  theme(axis.text = element_text(size = 10))
```


## Cell cycle effect

Assign each cell a score, based on its expression of G2/M and S phase markers as described in the Seurat workflow [here](https://satijalab.org/seurat/articles/cell_cycle_vignette.html).

```{r, warning=FALSE}
s.genes <- cc.genes.updated.2019$s.genes
g2m.genes <- cc.genes.updated.2019$g2m.genes

seu <- CellCycleScoring(seu, s.features = s.genes, g2m.features = g2m.genes, 
                        set.ident = TRUE)
```

PCA of cell cycle genes.

```{r, fig.asp=1.25}
DimPlot(seu, group.by = "Phase") -> p1

seu %>%
  RunPCA(features = c(s.genes, g2m.genes),
                      dims = 1:30, verbose = FALSE) %>%
  DimPlot(reduction = "pca") -> p2

(p2 / p1) + plot_layout(guides = "collect")
```

Distribution of cell cycle markers.

```{r, fig.asp=0.8}
# Visualize the distribution of cell cycle markers across
RidgePlot(seu, features = c("PCNA", "TOP2A", "MCM6", "MKI67"), ncol = 2,
          log = TRUE)
```

Using the `Seurat` *Alternate Workflow* from [here](https://satijalab.org/seurat/articles/cell_cycle_vignette.html), calculate the difference between the G2M and S phase scores so that signals separating non-cycling cells and cycling cells will be maintained, but differences in cell cycle phase among proliferating cells (which are often uninteresting), can be regressed out of the data.

```{r}
seu$CC.Difference <- seu$S.Score - seu$G2M.Score
```


## Integrate RNA data

Split by *batch* for integration. Normalise with `SCTransform`. Increase the strength of alignment by increasing `k.anchor` parameter to 20 as recommended in Seurat Fast integration with RPCA [vignette](https://satijalab.org/seurat/articles/integration_rpca.html).

First, integrate the RNA data.

```{r}
out <- here("data",
            "C133_Neeland_merged",
            glue("C133_Neeland_full_clean{ambient}_integrated_macrophages.SEU.rds"))

if(!file.exists(out)){
  DefaultAssay(seu) <- "RNA"
  VariableFeatures(seu) <- NULL
  seu[["pca"]] <- NULL
  seu[["umap"]] <- NULL
  
  seuLst <- SplitObject(seu, split.by = "Batch")
  rm(seu)
  gc()
  
  # normalise with SCTransform and regress out cell cycle score difference
  seuLst <- lapply(X = seuLst, FUN = SCTransform, method = "glmGamPoi",
                   vars.to.regress = "CC.Difference")
  # integrate RNA data
  features <- SelectIntegrationFeatures(object.list = seuLst,
                                        nfeatures = 3000)
  seuLst <- PrepSCTIntegration(object.list = seuLst, anchor.features = features)
  seuLst <- lapply(X = seuLst, FUN = RunPCA, features = features)
  anchors <- FindIntegrationAnchors(object.list = seuLst,
                                    normalization.method = "SCT",
                                    anchor.features = features,
                                    #k.anchor = 20,
                                    dims = 1:30, reduction = "rpca")
  seu <- IntegrateData(anchorset = anchors, 
                       #k.weight = min(100, min(sapply(seuLst, ncol)) - 5),
                       normalization.method = "SCT",
                       dims = 1:30)
  
  DefaultAssay(seu) <- "integrated"
  seu <- RunPCA(seu, dims = 1:30, verbose = FALSE) %>%
    RunUMAP(dims = 1:30, verbose = FALSE)

  saveRDS(seu, file = out)
  fs::file_chmod(out, "664")
  if(any(str_detect(fs::group_ids()$group_name, 
                    "oshlack_lab"))) fs::file_chown(out, 
                                                    group_id = "oshlack_lab")
  
} else {
  seu <- readRDS(file = out)
  
}
```


# Cluster data 

Perform clustering only on data that has ADT i.e. exclude batch 0.

## Dimensionality reduction (RNA)

Exclude any mitochondrial, ribosomal, immunoglobulin and HLA genes from variable genes list, to encourage clustering by cell type.

```{r, fig.width=9, fig.height=27}
gns <- AnnotationDbi::mapIds(org.Hs.eg.db,
                             keys = rownames(seu),
                             column = c("CHR"),
                             keytype = "SYMBOL",
                             multiVals = "first")
sex_genes <- names(gns[gns %in% c("X","Y")])
  
# remove HLA, immunoglobulin, MT, RP, MRP and sex genes from variable genes list
var_regex = '^HLA-|^IG[HJKL]|^MT-|^RPL|^MRPL' 
hvg <- VariableFeatures(seu)[!(grepl(var_regex, VariableFeatures(seu)) | 
                               (VariableFeatures(seu) %in% sex_genes))]
# assign edited variable gene list back to object 
VariableFeatures(seu) <- hvg

# redo PCA and UMAP 
seu <- RunPCA(seu, dims = 1:30, verbose = FALSE) %>%
    RunUMAP(dims = 1:30, verbose = FALSE)

DimHeatmap(seu, dims = 1:30, cells = 500, balanced = TRUE,
           reduction = "pca", assays = "integrated")
```

```{r}
ElbowPlot(seu, ndims = 30, reduction = "pca")
```

## Run clustering

Perform clustering at a range of resolutions and visualise to see which is appropriate to proceed with.

```{r, warning=FALSE, message=FALSE, fig.asp=1.25}
out <- here("data",
            "C133_Neeland_merged",
            glue("C133_Neeland_full_clean{ambient}_integrated_clustered_macrophages.SEU.rds"))

if(!file.exists(out)){
  DefaultAssay(seu) <- "integrated"
  seu <- FindNeighbors(seu, reduction = "pca", dims = 1:30)
  seu <- FindClusters(seu, algorithm = 3, 
                           resolution = seq(0.1, 1, by = 0.1))
  saveRDS(seu, file = out)
  fs::file_chmod(out, "664")
  if(any(str_detect(fs::group_ids()$group_name, 
                    "oshlack_lab"))) fs::file_chown(out, 
                                                    group_id = "oshlack_lab")
  
} else {
  seu <- readRDS(file = out)
  
}

clustree::clustree(seu, prefix = "integrated_snn_res.")
```

## View clusters

Choose most appropriate resolution based on `clustree` plot above.

```{r, fig.asp=1}
grp <- "integrated_snn_res.0.6"
# change factor ordering
seu@meta.data[,grp] <- fct_inseq(seu@meta.data[,grp])

DimPlot(seu, group.by = grp, label = T) + 
  theme(legend.position = "bottom")
```

```{r, fig.asp=0.8}
DimPlot(seu, reduction = "umap", group.by = "Phase", 
             label = FALSE, label.size = 3)
DimPlot(seu, reduction = "umap", group.by = "Disease", 
             label = FALSE, label.size = 3)
```


## Examine combined clusters

Number of cells per cluster.

```{r}
seu@meta.data %>%
  ggplot(aes(x = !!sym(grp), fill = !!sym(grp))) +
  geom_bar() +
  geom_text(aes(label = ..count..), stat = "count",
            vjust = -0.5, colour = "black", size = 2) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  NoLegend()
```

Visualise quality metrics by cluster. 

```{r, fig.asp=0.8}
seu@meta.data %>%
  ggplot(aes(x = !!sym(grp),
             y = nCount_RNA,
             fill = !!sym(grp))) +
  geom_violin(scale = "area") +
  scale_y_log10() +
  NoLegend() -> p2

seu@meta.data %>%
  ggplot(aes(x = !!sym(grp),
             y = nFeature_RNA,
             fill = !!sym(grp))) +
  geom_violin(scale = "area") +
  scale_y_log10() +
  NoLegend() -> p3

(p2 / p3) & theme(text = element_text(size = 8))
```

Check the batch composition of each of the clusters. 

```{r}
dittoBarPlot(seu,
             var = "Batch", 
             group.by = grp)
```

Check the sample compositions of clusters.

```{r, fig.asp=1}
dittoBarPlot(seu,
             var = "sample.id", 
             group.by = grp) + ggtitle("Samples") +
  theme(legend.position = "bottom")
```

Check the Azimuth labels compositions of clusters.

```{r, fig.width=9}
dittoBarPlot(seu,
             var = "predicted.ann_finest_level", 
             group.by = grp) + ggtitle("Samples") +
  theme(legend.position = "bottom")
```

Cluster composition by disease group.

```{r, fig.width=9}
dittoBarPlot(seu,
             var = "Disease", 
             group.by = grp) + ggtitle("Samples") +
  theme(legend.position = "bottom")
```

# RNA marker gene analysis

Adapted from Dr. Belinda Phipson's [work](https://bphipson.github.io/Human_Development_snRNAseq/14-MarkerAnalysisBCT.html) for [@Sim2021-cg].

## Test for marker genes using `limma`

```{r}
# limma-trend for DE
Idents(seu) <- grp

logcounts <- normCounts(DGEList(as.matrix(seu[["RNA"]]@counts)),
                        log = TRUE, prior.count = 0.5)
entrez <- AnnotationDbi::mapIds(org.Hs.eg.db,
                                keys = rownames(logcounts),
                                column = c("ENTREZID"),
                                keytype = "SYMBOL",
                                multiVals = "first")
# remove genes without entrez IDs as these are difficult to interpret biologically
logcounts <- logcounts[!is.na(entrez),]
# remove confounding genes from counts table e.g. mitochondrial, ribosomal etc.
logcounts <- logcounts[!str_detect(rownames(logcounts), var_regex),]

maxclust <- length(levels(Idents(seu))) - 1

clustgrp <- paste0("c", Idents(seu))
clustgrp <- factor(clustgrp, levels = paste0("c", 0:maxclust))
donor <- factor(seu$sample.id)

design <- model.matrix(~ 0 + clustgrp + donor)
colnames(design)[1:(length(levels(clustgrp)))] <- levels(clustgrp)

# Create contrast matrix
mycont <- matrix(NA, ncol = length(levels(clustgrp)),
                 nrow = length(levels(clustgrp)))
rownames(mycont) <- colnames(mycont) <- levels(clustgrp)
diag(mycont) <- 1
mycont[upper.tri(mycont)] <- -1/(length(levels(factor(clustgrp))) - 1)
mycont[lower.tri(mycont)] <- -1/(length(levels(factor(clustgrp))) - 1)

# Fill out remaining rows with 0s
zero.rows <- matrix(0, ncol = length(levels(clustgrp)),
                    nrow = (ncol(design) - length(levels(clustgrp))))
fullcont <- rbind(mycont, zero.rows)
rownames(fullcont) <- colnames(design)

fit <- lmFit(logcounts, design)

fit.cont <- contrasts.fit(fit, contrasts = fullcont)
fit.cont <- eBayes(fit.cont, trend = TRUE, robust = TRUE)

summary(decideTests(fit.cont))
```

Test relative to a threshold (TREAT).

```{r}
tr <- treat(fit.cont, lfc = 0.5)
dt <- decideTests(tr)
summary(dt)
```

Mean-difference (MD) plots per cluster. 

```{r, fig.asp=1}
par(mfrow=c(4,3))
par(mar=c(2,3,1,2))

for(i in 1:ncol(mycont)){
  plotMD(tr, coef = i, status = dt[,i], hl.cex = 0.5)
  abline(h = 0, col = "lightgrey")
  lines(lowess(tr$Amean, tr$coefficients[,i]), lwd = 1.5, col = 4)
}
```

## `limma` marker gene dotplot

```{r, fig.asp=2.25}
DefaultAssay(seu) <- "RNA"
contnames <- colnames(mycont)
top_markers <- NULL
n_markers <- 10

for(i in 1:ncol(mycont)){
  top <- topTreat(tr, coef = i, n = Inf)
  top <- top[top$logFC > 0, ]
  top_markers <- c(top_markers, 
                   setNames(rownames(top)[1:n_markers], 
                            rep(contnames[i], n_markers)))
}

top_markers <- top_markers[!is.na(top_markers)]
top_markers <- top_markers[!duplicated(top_markers)]
cols <- paletteer::paletteer_d("pals::glasbey")[factor(names(top_markers))]

DotPlot(seu,    
        features = unname(top_markers),
        group.by = grp,
        cols = c("azure1", "blueviolet"),
        dot.scale = 3, assay = "SCT") +
    RotatedAxis() +
    FontSize(y.text = 8, x.text = 12) +
    labs(y = element_blank(), x = element_blank()) +
    coord_flip() +
  theme(axis.text.y = element_text(color = cols)) +
  ggtitle("Top 10 cluster marker genes (no duplicates)")
```

## Save marker genes and pathways

The Broad MSigDB Reactome pathways are tested for each contrast using `cameraPR` from *limma*. The `cameraPR` method tests whether a set of genes is highly ranked relative to other genes in terms of differential expression, accounting for inter-gene correlation.

Prepare gene sets of interest.

```{r}
if(!file.exists(here("data/Hs.c2.cp.reactome.v7.1.entrez.rds")))
  download.file("https://bioinf.wehi.edu.au/MSigDB/v7.1/Hs.c2.cp.reactome.v7.1.entrez.rds",
                here("data/Hs.c2.cp.reactome.v7.1.entrez.rds"))

Hs.c2.reactome <- readRDS(here("data/Hs.c2.cp.reactome.v7.1.entrez.rds"))

gns <- AnnotationDbi::mapIds(org.Hs.eg.db, 
                             keys = rownames(tr), 
                             column = c("ENTREZID"),
                             keytype = "SYMBOL",
                             multiVals = "first")
```

Run pathway analysis and save results to file.

```{r}
options(scipen=-1, digits = 6)
contnames <- colnames(mycont)
dirName <- here("output",
                "cluster_markers",
                glue("RNA{ambient}"),
                "macrophages")

if(!dir.exists(dirName)) dir.create(dirName, recursive = TRUE)

for(c in colnames(tr)){
  top <- topTreat(tr, coef = c, n = Inf)
  top <- top[top$logFC > 0, ]

  write.csv(top[1:100, ] %>%
              rownames_to_column(var = "Symbol"),
            file = glue("{dirName}/up-cluster-limma-{c}.csv"),
            sep = ",",
            quote = FALSE,
            col.names = NA,
            row.names = TRUE)

  # get marker indices
  c2.id <- ids2indices(Hs.c2.reactome, unname(gns[rownames(tr)]))
  # gene set testing results
  cameraPR(tr$t[,glue("{c}")], c2.id) %>%
    rownames_to_column(var = "Pathway") %>%
    dplyr::filter(Direction == "Up") %>%
    slice_head(n = 50) %>%
    write.csv(file = here(glue("{dirName}/REACTOME-cluster-limma-{c}.csv")),
            sep = ",",
            quote = FALSE,
            col.names = NA,
            row.names = TRUE)
}
```


# ADT marker analysis
## Find all marker ADT using `limma`

```{r}
# identify isotype controls for DSB ADT normalisation
read_csv(file = here("data",
                     "C133_Neeland_batch1",
                     "data",
                     "sample_sheets",
                     "ADT_features.csv")) -> adt_data

pattern <- "anti-human/mouse |anti-human/mouse/rat |anti-mouse/human "
adt_data$name <- gsub(pattern, "", adt_data$name)
# change ADT rownames to antibody names
DefaultAssay(seu) <- "ADT"
if(all(rownames(seu[["ADT"]]@counts) == adt_data$id)){
  adt <- seu[["ADT"]]@counts
  rownames(adt) <- adt_data$name
}

adt_data %>%
  dplyr::filter(grepl("[Ii]sotype", name)) %>%
  pull(name) -> isotype_controls
  
# normalise ADT using DSB normalisation
adt_dsb <- ModelNegativeADTnorm(cell_protein_matrix = adt,
                                denoise.counts = TRUE,
                                use.isotype.control = TRUE,
                                isotype.control.name.vec = isotype_controls)
```

Running the `limma` analysis on the normalised counts.

```{r}
# limma-trend for DE
Idents(seu) <- grp

logcounts <- adt_dsb
# remove isotype controls from marker analysis
logcounts <- logcounts[!rownames(logcounts) %in% isotype_controls,]
maxclust <- length(levels(Idents(seu))) - 1

clustgrp <- paste0("c", Idents(seu))
clustgrp <- factor(clustgrp, levels = paste0("c", 0:maxclust))
donor <- seu$sample.id

design <- model.matrix(~ 0 + clustgrp + donor)
colnames(design)[1:(length(levels(clustgrp)))] <- levels(clustgrp)

# Create contrast matrix
mycont <- matrix(NA, ncol = length(levels(clustgrp)),
                 nrow = length(levels(clustgrp)))
rownames(mycont) <- colnames(mycont) <- levels(clustgrp)
diag(mycont) <- 1
mycont[upper.tri(mycont)] <- -1/(length(levels(factor(clustgrp))) - 1)
mycont[lower.tri(mycont)] <- -1/(length(levels(factor(clustgrp))) - 1)

# Fill out remaining rows with 0s
zero.rows <- matrix(0, ncol = length(levels(clustgrp)),
                    nrow = (ncol(design) - length(levels(clustgrp))))
fullcont <- rbind(mycont, zero.rows)
rownames(fullcont) <- colnames(design)

fit <- lmFit(logcounts, design)
fit.cont <- contrasts.fit(fit, contrasts = fullcont)
fit.cont <- eBayes(fit.cont, trend = TRUE, robust = TRUE)

summary(decideTests(fit.cont))
```

Test relative to a threshold (TREAT).

```{r}
tr <- treat(fit.cont, lfc = 0.1)
dt <- decideTests(tr)
summary(dt)
```

## ADT marker dot plot 

Dot plot of the top 5 ADT markers per cluster without duplication.

```{r, fig.asp=1.2, fig.width=8}
contnames <- colnames(mycont)
top_markers <- NULL
n_markers <- 5

for (i in 1:length(contnames)){
  top <- topTreat(tr, coef = i, n = Inf, p.value = 0.05)
  if(nrow(top) > 0){
    top <- top[top$logFC > 0,]
    top_markers <- c(top_markers, 
                     setNames(rownames(top)[1:min(n_markers, nrow(top))], 
                              rep(contnames[i], min(n_markers, nrow(top)))))
  }
}

top_markers <- top_markers[!is.na(top_markers)]
top_markers <- top_markers[!duplicated(top_markers)]
cols <- paletteer::paletteer_d("pals::glasbey")[factor(names(top_markers))][!duplicated(top_markers)]

# add DSB normalised data to Seurat assay for plotting
seu[["ADT.dsb"]] <- CreateAssayObject(data = logcounts)
DotPlot(seu, 
        group.by = grp,
        features = unname(top_markers), 
        cols = c("azure1", "blueviolet"),
        assay = "ADT.dsb")  +
  RotatedAxis() + 
  FontSize(y.text = 8, x.text = 9) +
  labs(y = element_blank(), x = element_blank()) +
  theme(axis.text.y = element_text(color = cols),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10)) +
    coord_flip() +
  ggtitle("Top 5 cluster markers ADTs (no duplicates)")
```

## ADT marker heatmap

Make data frame of proteins, clusters, expression levels.

```{r}
cbind(seu@meta.data %>%
        dplyr::select(!!sym(grp)),
      as.data.frame(t(seu@assays$ADT.dsb@data))) %>%
  rownames_to_column(var = "cell") %>%
  pivot_longer(c(-!!sym(grp), -cell), 
               names_to = "ADT",
               values_to = "expression") %>%
  dplyr::group_by(!!sym(grp), ADT) %>%
  dplyr::summarize(Expression = mean(expression)) %>%
  ungroup() -> dat

# plot expression density to select heatmap colour scale range
plot(density(dat$Expression))
```

```{r, fig.asp=1.2, fig.width=8}
dat %>%
  dplyr::filter(ADT %in% top_markers) |>
  heatmap(
    .column = !!sym(grp),
    .row = ADT,
    .value = Expression,
    row_order = top_markers, 
    scale = "none",
    rect_gp = grid::gpar(col = "white", lwd = 1),
    show_row_names = TRUE,
    cluster_columns = FALSE,
    cluster_rows = FALSE,
    column_names_gp = grid::gpar(fontsize = 10),
    column_title_gp = grid::gpar(fontsize = 12),
    row_names_gp = grid::gpar(fontsize = 8, col = cols[order(top_markers)]),
    row_title_gp = grid::gpar(fontsize = 12),
    column_title_side = "top",
    palette_value = circlize::colorRamp2(seq(0, 3, length.out = 256),
                                         viridis::magma(256)),
    heatmap_legend_param = list(direction = "vertical"))
```

## Save ADT markers

```{r}
options(scipen=-1, digits = 6)
contnames <- colnames(mycont)
dirName <- here("output",
                "cluster_markers",
                glue("ADT{ambient}"),
                "macrophages")
if(!dir.exists(dirName)) dir.create(dirName, recursive = TRUE)

for(c in contnames){
  top <- topTreat(tr, coef = c, n = Inf)
  top <- top[top$logFC > 0, ]
  write.csv(top,
            file = glue("{dirName}/up-cluster-limma-{c}.csv"),
            sep = ",",
            quote = FALSE,
            col.names = NA,
            row.names = TRUE)
}
```


# Session info










