---
title: "Annotate T/NK clusters"
author: "Jovana Maksimovic"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
site: workflowr::wflow_site
output:
  bookdown::html_document2:
    base_format: workflowr::wflow_html
    toc: yes
    toc_float: yes
    theme: cosmo
    highlight: textmate
    number_sections: true
---

```{r, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

# Load libraries

```{r, message=FALSE, echo=FALSE}
suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(edgeR)
  library(tidyverse)
  library(ggplot2)
  library(Seurat)
  library(dittoSeq)
  library(here)
  library(glue)
  library(patchwork)
  library(paletteer)
  library(tidyHeatmap)
  library(readxl)
  library(AnnotationDbi)
  library(org.Hs.eg.db)
  library(speckle)
  library(ggh4x)
  library(dsb)
})

set.seed(42)
options(scipen=999)
options(future.globals.maxSize = 6500 * 1024^2)
```

# Load Data

```{r, warning=FALSE, message=FALSE}
ambient <- "_decontx"
out <- here("data",
            "C133_Neeland_merged",
            glue("C133_Neeland_full_clean{ambient}_integrated_clustered_mapped_t_cells.ADT.SEU.rds"))
seuInt <- readRDS(file = out)

seuInt
```
## Update group labels

```{r}
seuInt@meta.data %>%
  data.frame %>%
  mutate(Status = ifelse(str_detect(Treatment, "ivacaftor"),
                         "CF ivacaftor",
                         ifelse(str_detect(Treatment, "orkambi"),
                                "CF lumacaftor-ivacaftor",
                                ifelse(Treatment == "untreated",
                                       "CF no-modulator",
                                       "non-CF control"))),
         Status_sub = ifelse(str_detect(Treatment, "ivacaftor"),
                         "CF.IVA",
                         ifelse(str_detect(Treatment, "orkambi"),
                                "CF.LUMA_IVA",
                                ifelse(Treatment == "untreated",
                                       "CF.NO_MOD",
                                       "NON_CF.CTRL"))),
         Group = ifelse(!Status_sub %in% "NON_CF.CTRL", 
                        paste(Status_sub, 
                              toupper(substr(Severity, 1, 1)),
                              sep = "."), 
                        Status_sub),
         Severity = tolower(Severity),
         Participant = strsplit2(sample.id, ".", fixed = TRUE)[,1]) -> seuInt@meta.data
```


# Sub-cluster labelling
## Load manual annotations

```{r}
labels <- read_excel(here("data",
                          "cluster_annotations",
                          "T-NK_ambientRNAremoval_21.03.24.xlsx"),
                     skip = 1)

# set selected cluster resolution
grp <- "wsnn_res.0.6"
seuInt@meta.data %>%
  rownames_to_column(var = "cell") %>%
  left_join(labels %>%
              mutate(Cluster = as.factor(Cluster),
                     Annotation = as.factor(Annotation),
                     Broad = as.factor(Broad)),
            by = c("wsnn_res.0.6" = "Cluster")) %>%
  column_to_rownames(var = "cell") -> seuInt@meta.data

seuInt$Annotation <- fct_drop(seuInt$Annotation)
seuInt$Broad <- fct_drop(seuInt$Broad)
seuInt
```

## Visualise annotations

```{r, fig.asp=0.8}
options(ggrepel.max.overlaps = Inf)
DimPlot(seuInt, reduction = 'umap', label = TRUE, repel = TRUE, 
        label.size = 3, group.by = grp) + 
  NoLegend() -> p1

cluster_pal <- "ggsci::category20_d3"
DimPlot(seuInt, reduction = 'umap', label = FALSE, group.by = "Annotation") + 
  scale_color_paletteer_d(cluster_pal) +
  theme(text = element_text(size = 9),
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  NoLegend() -> p2

p1
LabelClusters(p2, id = "Annotation", repel = TRUE, 
              size = 2.5, box = TRUE, fontfamily = "arial")
```

### No. cells per cluster

```{r, fig.asp=1}
seuInt@meta.data %>%
  ggplot(aes(x = Annotation, fill = Annotation)) +
  geom_bar() +
  geom_text(aes(label = after_stat(count)), stat = "count",
            vjust = -0.5, colour = "black", size = 2) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  NoLegend() +
  scale_fill_paletteer_d(cluster_pal)
```



```{r}
seuInt@meta.data %>% 
  dplyr::count(Annotation) %>% 
  mutate(perc = round(n/sum(n)*100, 1)) %>%
  dplyr::rename(`Cell Label` = "Annotation", 
                `No. Cells` = n,
                `% Cells` = perc) %>%
  knitr::kable()
```

# RNA marker gene analysis

Adapted from Dr. Belinda Phipson's [work](https://bphipson.github.io/Human_Development_snRNAseq/14-MarkerAnalysisBCT.html) for [@Sim2021-cg].

## Test for marker genes using `limma`

```{r}
# limma-trend for DE
Idents(seuInt) <- "Annotation"

out <- here("data",
            "C133_Neeland_merged",
            glue("C133_Neeland_full_clean{ambient}_TNK_cells_logcounts.SEU.rds"))

if(!file.exists(out)){
  logcounts <- normCounts(DGEList(as.matrix(seuInt[["RNA"]]@counts)),
                          log = TRUE, prior.count = 0.5)
  entrez <- AnnotationDbi::mapIds(org.Hs.eg.db,
                                  keys = rownames(logcounts),
                                  column = c("ENTREZID"),
                                  keytype = "SYMBOL",
                                  multiVals = "first")
  # remove genes without entrez IDs as these are difficult to interpret biologically
  logcounts <- logcounts[!is.na(entrez),]
  saveRDS(logcounts, file = out)
  
} else {
  logcounts <- readRDS(out)  
}

maxclust <- length(levels(Idents(seuInt))) - 1

clustgrp <- seuInt$Annotation
clustgrp <- factor(clustgrp)
donor <- factor(seuInt$sample.id)
batch <- factor(seuInt$Batch)

design <- model.matrix(~ 0 + clustgrp + donor)
colnames(design)[1:(length(levels(clustgrp)))] <- levels(clustgrp)

# Create contrast matrix
mycont <- matrix(NA, ncol = length(levels(clustgrp)),
                 nrow = length(levels(clustgrp)))
rownames(mycont) <- colnames(mycont) <- levels(clustgrp)
diag(mycont) <- 1
mycont[upper.tri(mycont)] <- -1/(length(levels(factor(clustgrp))) - 1)
mycont[lower.tri(mycont)] <- -1/(length(levels(factor(clustgrp))) - 1)

# Fill out remaining rows with 0s
zero.rows <- matrix(0, ncol = length(levels(clustgrp)),
                    nrow = (ncol(design) - length(levels(clustgrp))))
fullcont <- rbind(mycont, zero.rows)
rownames(fullcont) <- colnames(design)

fit <- lmFit(logcounts, design)

fit.cont <- contrasts.fit(fit, contrasts = fullcont)
fit.cont <- eBayes(fit.cont, trend = TRUE, robust = TRUE)

summary(decideTests(fit.cont))
```

Test relative to a threshold (TREAT).

```{r}
tr <- treat(fit.cont, lfc = 0.25)
dt <- decideTests(tr)
summary(dt)
```

Mean-difference (MD) plots per cluster. 

```{r, fig.asp=1}
par(mfrow=c(4,3))
par(mar=c(2,3,1,2))

for(i in 1:ncol(mycont)){
  plotMD(tr, coef = i, status = dt[,i], hl.cex = 0.5)
  abline(h = 0, col = "lightgrey")
  lines(lowess(tr$Amean, tr$coefficients[,i]), lwd = 1.5, col = 4)
}
```

## `limma` marker gene dotplot

```{r, fig.width=12, fig.asp=0.4}
DefaultAssay(seuInt) <- "RNA"
contnames <- colnames(mycont)
top_markers <- NULL
n_markers <- 5

for(i in 1:ncol(mycont)){
  top <- topTreat(tr, coef = i, n = Inf)
  top <- top[top$logFC > 0, ]
  top_markers <- c(top_markers, 
                   setNames(rownames(top)[1:n_markers], 
                            rep(contnames[i], n_markers)))
}

top_markers <- top_markers[!is.na(top_markers)]
d <- duplicated(top_markers)
top_markers <- top_markers[!d]
geneCols <- paletteer_d(cluster_pal)[factor(names(top_markers))]

strip <- strip_themed(background_x = elem_list_rect(fill = unique(geneCols)))
DotPlot(seuInt,
        features = unname(top_markers),
        group.by = "Annotation",
        cols = c("azure1", "blueviolet"),
        dot.scale = 3,
        assay = "SCT") +
  FontSize(x.text = 9, y.text = 9) +
  labs(y = element_blank(), x = element_blank()) +
  facet_grid2(~names(top_markers), 
              scales = "free_x", 
              space = "free_x", 
              strip = strip) +
  theme(axis.text.x = element_text(angle = 90,
                                   hjust = 1,
                                   vjust = 0.5),
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 9),
        strip.text = element_text(size = 0),
        text = element_text(family = "arial"),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        panel.spacing = unit(2, "mm")) 
```

## Test for marker genes using `Seurat`

```{r}
DefaultAssay(seuInt) <- "RNA"
Idents(seuInt) <- "Annotation"

out <- here("data/cluster_annotations/seurat_markers_TNK_cells.rds")

if(!file.exists(out)){
  # restrict genes to same set as for limma analysis
  markers <- FindAllMarkers(seuInt, only.pos = TRUE, 
                            features = rownames(logcounts))
  saveRDS(markers, file = out)
  
} else {
  markers <- readRDS(out)
  
}

head(markers) %>% knitr::kable()
```

## `Seurat` marker gene dotplot

```{r, fig.width=13, fig.asp=0.4}
DefaultAssay(seuInt) <- "RNA"

maxGenes <- 5
markers %>%
  group_by(cluster) %>%
  top_n(n = maxGenes, wt = avg_log2FC) -> top5

sig <- top5$gene
d <- duplicated(sig)
geneCols <- paletteer_d(cluster_pal)[top5$cluster][!d]

strip <- strip_themed(background_x = elem_list_rect(fill = unique(geneCols)))
DotPlot(seuInt,
        features = sig[!d],
        group.by = "Annotation",
        cols = c("azure1", "blueviolet"),
        dot.scale = 3,
        assay = "SCT") +
  FontSize(x.text = 9, y.text = 9) +
  labs(y = element_blank(), x = element_blank()) +
  facet_grid2(~top5$cluster[!d], 
              scales = "free_x", 
              space = "free_x", 
              strip = strip) +
  theme(axis.text.x = element_text(angle = 90,
                                   hjust = 1,
                                   vjust = 0.5),
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 9),
        strip.text = element_text(size = 0),
        text = element_text(family = "arial"),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        panel.spacing = unit(2, "mm")) 
```

## Visualise ADTs

Make data frame of proteins, clusters, expression levels.

```{r}
out <- here("data",
            "C133_Neeland_merged",
            glue("C133_Neeland_full_clean{ambient}_TNK_cells_adt_dsb.SEU.rds"))

if(!file.exists(out)){
  read_csv(file = here("data",
                       "C133_Neeland_batch1",
                       "data",
                       "sample_sheets",
                       "ADT_features.csv")) -> adt_data
  
  pattern <- "anti-human/mouse |anti-human/mouse/rat |anti-mouse/human |anti-human "
  adt_data$name <- gsub(pattern, "", adt_data$name)
  adt <- seuInt[["ADT"]]@counts
  if(all(rownames(seuInt[["ADT"]]@counts) == adt_data$id)) rownames(adt) <- adt_data$name
  adt_data %>%
  dplyr::filter(grepl("[Ii]sotype", name)) %>%
  pull(name) -> isotype_controls
  
  # normalise ADT using DSB normalisation
  adt_dsb <- ModelNegativeADTnorm(cell_protein_matrix = adt,
                                  denoise.counts = TRUE,
                                  use.isotype.control = TRUE,
                                  isotype.control.name.vec = isotype_controls)
  saveRDS(adt_dsb, file = out)
  
} else {
  adt_dsb <- readRDS(out)
  
}

m <- match(colnames(seuInt), colnames(adt_dsb)) # remove cells not present in Seurat obj
seuInt[["ADT.dsb"]] <- CreateAssayObject(data = adt_dsb[,m])
```


```{r, fig.asp=0.6, fig.width=6}
ADTs <- read_csv(file = here("data",
                       "Proteins_T-NK_22.04.22.csv"))
pattern <- "anti-human/mouse |anti-human/mouse/rat |anti-mouse/human |anti-human "
ADTs$Description <- gsub(pattern, "", ADTs$Description)

DotPlot(seuInt,
        features = ADTs$Description,
        group.by = "Annotation",
        cols = c("azure1", "blueviolet"),
        dot.scale = 2.5,
        assay = "ADT.dsb") +
  FontSize(x.text = 9, y.text = 9) +
  labs(y = element_blank(), x = element_blank()) +
  theme(axis.text.x = element_text(angle = 90,
                                   hjust = 1,
                                   vjust = 0.5),
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 9),
        strip.text = element_text(size = 0),
        text = element_text(family = "arial"),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        panel.spacing = unit(2, "mm")) 
```


# Save data

```{r}
out <- here("data",
            "C133_Neeland_merged",
            glue("C133_Neeland_full_clean{ambient}_t_cells_annotated_diet.SEU.rds"))
if(!file.exists(out)){
  DefaultAssay(seuInt) <- "RNA"
  saveRDS(DietSeurat(seuInt, assays = "RNA"), out)
}

out <- here("data",
            "C133_Neeland_merged",
            glue("C133_Neeland_full_clean{ambient}_t_cells_annotated_full.SEU.rds"))
if(!file.exists(out)){
  DefaultAssay(seuInt) <- "RNA"
  saveRDS(seuInt, out)
}
```

<!-- # Panel figures -->

<!-- ```{r, fig.width=10, fig.asp=1.3} -->
<!-- layout = "AAAABB -->
<!--           AAAABB -->
<!--           AAAABB -->
<!--           CCCCCC -->
<!--           CCCCCC -->
<!--           DDDDDD -->
<!--           DDDDDD -->
<!--           DDDDDD" -->
<!-- ((f2a + ggtitle("")) +  -->
<!--    f2b +  -->
<!--    f2c +  -->
<!--    wrap_heatmap(f2d)) +  -->
<!--   plot_layout(design = layout) + -->
<!--   plot_annotation(tag_levels = "A") & -->
<!--   theme(plot.tag = element_text(size = 14, face = "bold")) -->
<!-- ``` -->

# Session info {.appendix}

