---
title: "Inflammation of Paediatric Pulmonary Diseases"
subtitle: "Cell type proportions analysis - annotation level 1"
author: "Jovana Maksimovic"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
site: workflowr::wflow_site
output:
  bookdown::html_document2:
    base_format: workflowr::wflow_html
    toc: yes
    toc_float: yes
    theme: cosmo
    highlight: textmate
    number_sections: true
---

```{r, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

# Load libraries

```{r, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(edgeR)
  library(tidyverse)
  library(ggplot2)
  library(Seurat)
  library(glmGamPoi)
  library(dittoSeq)
  library(clustree)
  library(AnnotationDbi)
  library(org.Hs.eg.db)
  library(glue)
  library(speckle)
  library(patchwork)
  library(paletteer)
  library(tidyHeatmap)
  library(here)
})

set.seed(42)
options(scipen=999)
options(future.globals.maxSize = 6500 * 1024^2)
```

# Load Data

```{r}
files <- list.files(here("data/C133_Neeland_merged"),
                    pattern = "C133_Neeland_full_clean.*(macrophages|t_cells|other_cells)_annotated_diet.SEU.rds",
                    full.names = TRUE)

seuLst <- lapply(files[2:4], function(f) readRDS(f))

seu <- merge(seuLst[[1]], 
             y = c(seuLst[[2]], 
                   seuLst[[3]]))
seu
```

```{r, echo=FALSE}
# cleanup obsolete objects
rm(seuLst)
gc()
```


# Analyse Cell type proportions 

```{r}
# Differences in cell type proportions
props <- getTransformedProps(clusters = seu$ann_level_1,
                             sample = seu$sample.id, transform="asin")
props$Proportions %>% knitr::kable()
```


## Cell type proportions by sample

```{r, fig.asp=0.6, fig.width=12}
props$Proportions %>%
  data.frame %>%
  inner_join(seu@meta.data %>%
               dplyr::select(sample.id, 
                             Disease,
                             Treatment,
                             Severity,
                             Group,
                             Batch, 
                             Age, 
                             Sex), 
             by = c("sample" = "sample.id")) %>%
  distinct()-> dat

ggplot(dat, aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", color = "black", size = 0.1) +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5,
                                   hjust = 1),
        legend.text = element_text(size = 8),
        legend.position = "bottom") +
  labs(y = "Proportion", fill = "Cell Label") +
  scale_fill_paletteer_d("miscpalettes::pastel", direction = 1) +
  facet_grid(~Group, scales = "free_x", space = "free_x")
```

## No. cells per sample

```{r, fig.asp=0.5, fig.width=9}
props$Counts %>%
  data.frame %>%
  inner_join(seu@meta.data %>%
               dplyr::select(sample.id, 
                             ann_level_1,
                             Disease,
                             Treatment,
                             Severity, 
                             Group,
                             Group_severity,
                             Batch, 
                             Age, 
                             Sex), 
             by = c("sample" = "sample.id", "clusters" = "ann_level_1")) %>%
  distinct() -> dat

ggplot(dat, aes(x = sample, y = Freq, fill = Disease)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5,
                                   hjust = 1),
        legend.text = element_text(size = 8),
        legend.position = "bottom") +
  labs(y = "No. cells", fill = "Disease") +
  facet_grid(~Group, scales = "free_x", space = "free_x") +
  geom_hline(yintercept = 1000, linetype = "dashed")
```

<!-- Remove sample with too few cells. -->

<!-- ```{r} -->
<!-- seu <- subset(seu, cells = which(!seu$sample.id %in% c("sample_23.1"))) -->

<!-- # Differences in cell type proportions -->
<!-- props <- getTransformedProps(clusters = seu$ann_level_1, -->
<!--                              sample = seu$sample.id, transform="asin") -->
<!-- props$Proportions %>% knitr::kable() -->
<!-- ``` -->

## Cell proportions by sample

```{r, fig.asp=0.75, fig.width=9}
props$Proportions %>%
  data.frame %>%
  inner_join(seu@meta.data %>%
               dplyr::select(sample.id, 
                             ann_level_1,
                             Disease,
                             Treatment,
                             Severity, 
                             Group,
                             Group_severity,
                             Batch, 
                             Age, 
                             Sex), 
             by = c("sample" = "sample.id", "clusters" = "ann_level_1")) %>%
  distinct() -> dat

ggplot(dat, aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", color = "black", size = 0.1) +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5,
                                   hjust = 1),
        legend.text = element_text(size = 8),
        legend.position = "bottom") +
  labs(y = "Proportion", fill = "Cell Label") +
  scale_fill_paletteer_d("miscpalettes::pastel", direction = 1) +
  facet_grid(~Group, scales = "free_x", space = "free_x")
```

## Cell proportions by cell type

```{r}
props$Proportions %>%
  data.frame %>%
  inner_join(seu@meta.data %>%
               dplyr::select(sample.id, 
                             ann_level_1,
                             Disease,
                             Treatment,
                             Severity, 
                             Group,
                             Group_severity,
                             Batch, 
                             Age, 
                             Sex), 
             by = c("sample" = "sample.id", "clusters" = "ann_level_1")) %>%
  distinct() -> dat

ggplot(dat, 
       aes(x = clusters, y = Freq, fill = clusters)) +
  geom_boxplot(outlier.size = 0.1, size = 0.25) +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1),
        legend.text = element_text(size = 8)) +
  labs(y = "Proportion") +
  scale_fill_paletteer_d("miscpalettes::pastel", direction = 1) +
  NoLegend()
```


# Explore sources of variation
## Cell count data

Look at the sources of variation in the raw cell count level data.

```{r, fig.asp=1}
dims <- list(c(1,2), c(2:3), c(3,4), c(4,5))
p <- vector("list", length(dims))

for(i in 1:length(dims)){
  mds <- plotMDS(props$Counts,
    gene.selection = "common",
    plot = FALSE, dim.plot = dims[[i]])
  
  data.frame(x = mds$x, 
             y = mds$y,
             sample = rownames(mds$distance.matrix.squared)) %>%
    left_join(seu@meta.data %>%
               dplyr::select(sample.id, 
                             Disease,
                             Treatment,
                             Severity, 
                             Batch, 
                             Age, 
                             Sex), 
             by = c("sample" = "sample.id")) %>%
    distinct() -> dat
  
  p[[i]] <- ggplot(dat, aes(x = x, y = y, 
                            shape = as.factor(Disease),
                            color = as.factor(Batch))) +
    geom_point(size = 3) +
    labs(x = glue("Principal Component {dims[[i]][1]}"),
         y = glue("Principal Component {dims[[i]][2]}"),
         colour = "Batch",
         shape = "Disease") +
    theme(legend.direction = "horizontal",
          legend.text = element_text(size = 8),
          legend.title = element_text(size = 9),
          axis.text = element_text(size = 8),
          axis.title = element_text(size = 9)) 
}

wrap_plots(p, cols = 2) + plot_layout(guides = "collect") &
  theme(legend.position = "bottom") 
```

```{r, fig.asp=1}
dims <- list(c(1,2), c(2:3), c(3,4), c(4,5))
p <- vector("list", length(dims))

for(i in 1:length(dims)){
  mds <- plotMDS(props$Counts, 
                 gene.selection = "common",
                 plot = FALSE, dim.plot = dims[[i]])
  
   data.frame(x = mds$x, 
             y = mds$y,
             sample = rownames(mds$distance.matrix.squared)) %>%
    left_join(seu@meta.data %>%
               dplyr::select(sample.id, 
                             Disease,
                             Treatment,
                             Severity,
                             Participant,
                             Batch, 
                             Age, 
                             Sex), 
             by = c("sample" = "sample.id")) %>%
    group_by(sample) %>%
    mutate(ncells = n()) %>%
    ungroup() %>%
    distinct() -> dat
  
  p[[i]] <- ggplot(dat, aes(x = x, y = y, 
                            colour = log2(ncells)))+
    geom_text(aes(label = str_remove_all(sample, "sample_")), size = 2.5) +
    labs(x = glue("Principal Component {dims[[i]][1]}"),
         y = glue("Principal Component {dims[[i]][2]}"),
         colour = "Log2 No. Cells") +
    theme(legend.direction = "horizontal",
          legend.text = element_text(size = 8),
          legend.title = element_text(size = 9),
          axis.text = element_text(size = 8),
          axis.title = element_text(size = 9)) +
    scale_colour_viridis_c(option = "magma")
}

wrap_plots(p, cols = 2) + plot_layout(guides = "collect") &
  theme(legend.position = "bottom") 
```

## Cell proportion data

Look at the sources of variation in the cell proportions data.

```{r, fig.asp=1}
dims <- list(c(1,2), c(2:3), c(3,4), c(4,5))
p <- vector("list", length(dims))

for(i in 1:length(dims)){
  mds <- plotMDS(props$TransformedProps,
    gene.selection = "common",
    plot = FALSE, dim.plot = dims[[i]])
  
  data.frame(x = mds$x, 
             y = mds$y,
             sample = rownames(mds$distance.matrix.squared)) %>%
    left_join(seu@meta.data %>%
               dplyr::select(sample.id, 
                             Disease,
                             Treatment,
                             Severity, 
                             Batch, 
                             Age, 
                             Sex), 
             by = c("sample" = "sample.id")) %>%
    distinct() -> dat
  
  p[[i]] <- ggplot(dat, aes(x = x, y = y, 
                            shape = as.factor(Disease),
                            color = as.factor(Batch)))+
    geom_point(size = 3) +
    labs(x = glue("Principal Component {dims[[i]][1]}"),
         y = glue("Principal Component {dims[[i]][2]}"),
         colour = "Batch",
         shape = "Disease") +
    theme(legend.direction = "horizontal",
          legend.text = element_text(size = 8),
          legend.title = element_text(size = 9),
          axis.text = element_text(size = 8),
          axis.title = element_text(size = 9))
}

wrap_plots(p, cols = 2) + plot_layout(guides = "collect") &
  theme(legend.position = "bottom") 
```

```{r, fig.asp=1}
dims <- list(c(1,2), c(2:3), c(3,4), c(4,5))
p <- vector("list", length(dims))

for(i in 1:length(dims)){
  mds <- plotMDS(props$TransformedProps,
    gene.selection = "common",
    plot = FALSE, dim.plot = dims[[i]])
  
  data.frame(x = mds$x, 
             y = mds$y,
             sample = rownames(mds$distance.matrix.squared)) %>%
    left_join(seu@meta.data %>%
               dplyr::select(sample.id, 
                             Disease,
                             Treatment,
                             Severity, 
                             Batch, 
                             Age, 
                             Sex), 
             by = c("sample" = "sample.id")) %>%
    distinct() -> dat
  
  p[[i]] <- ggplot(dat, aes(x = x, y = y, 
                            shape = as.factor(Disease),
                            color = Sex))+
    geom_point(size = 3) +
    labs(x = glue("Principal Component {dims[[i]][1]}"),
         y = glue("Principal Component {dims[[i]][2]}"),
         colour = "Sex",
         shape = "Disease") +
    theme(legend.direction = "horizontal",
          legend.text = element_text(size = 8),
          legend.title = element_text(size = 9),
          axis.text = element_text(size = 8),
          axis.title = element_text(size = 9))
}

wrap_plots(p, cols = 2) + plot_layout(guides = "collect") &
  theme(legend.position = "bottom") 
```

```{r, fig.asp=1}
dims <- list(c(1,2), c(2:3), c(3,4), c(4,5))
p <- vector("list", length(dims))

for(i in 1:length(dims)){
  mds <- plotMDS(props$TransformedProps, 
                 gene.selection = "common",
                 plot = FALSE, dim.plot = dims[[i]])
  
   data.frame(x = mds$x, 
             y = mds$y,
             sample = rownames(mds$distance.matrix.squared)) %>%
    left_join(seu@meta.data %>%
               dplyr::select(sample.id, 
                             Disease,
                             Treatment,
                             Severity,
                             Participant,
                             Batch, 
                             Age, 
                             Sex), 
             by = c("sample" = "sample.id")) %>%
    group_by(sample) %>%
    mutate(ncells = n()) %>%
    ungroup() %>%
    distinct() -> dat
  
  p[[i]] <- ggplot(dat, aes(x = x, y = y, 
                            colour = log2(Age)))+
    geom_text(aes(label = str_remove_all(sample, "sample_")), size = 2.5) +
    labs(x = glue("Principal Component {dims[[i]][1]}"),
         y = glue("Principal Component {dims[[i]][2]}"),
         colour = "Log2 Age") +
    theme(legend.direction = "horizontal",
          legend.text = element_text(size = 8),
          legend.title = element_text(size = 9),
          axis.text = element_text(size = 8),
          axis.title = element_text(size = 9)) +
    scale_colour_viridis_c(option = "magma")
}

wrap_plots(p, cols = 2) + plot_layout(guides = "collect") &
  theme(legend.position = "bottom") 
```

```{r, fig.asp=1}
dims <- list(c(1,2), c(2:3), c(3,4), c(4,5))
p <- vector("list", length(dims))

for(i in 1:length(dims)){
  mds <- plotMDS(props$TransformedProps, 
                 gene.selection = "common",
                 plot = FALSE, dim.plot = dims[[i]])
  
   data.frame(x = mds$x, 
             y = mds$y,
             sample = rownames(mds$distance.matrix.squared)) %>%
    left_join(seu@meta.data %>%
               dplyr::select(sample.id, 
                             Disease,
                             Treatment,
                             Severity,
                             Participant,
                             Batch, 
                             Age, 
                             Sex), 
             by = c("sample" = "sample.id")) %>%
    group_by(sample) %>%
    mutate(ncells = n()) %>%
    ungroup() %>%
    distinct() -> dat
  
  p[[i]] <- ggplot(dat, aes(x = x, y = y, 
                            colour = log2(ncells)))+
    geom_text(aes(label = str_remove_all(sample, "sample_")), size = 2.5) +
    labs(x = glue("Principal Component {dims[[i]][1]}"),
         y = glue("Principal Component {dims[[i]][2]}"),
         colour = "Log2 No. Cells") +
    theme(legend.direction = "horizontal",
          legend.text = element_text(size = 8),
          legend.title = element_text(size = 9),
          axis.text = element_text(size = 8),
          axis.title = element_text(size = 9)) +
    scale_colour_viridis_c(option = "magma")
}

wrap_plots(p, cols = 2) + plot_layout(guides = "collect") &
  theme(legend.position = "bottom") 
```


# Principal components versus traits

Principal components analysis (PCA) allows us to mathematically determine the sources of variation in the data. We can then investigate whether these correlate with any of the specifed covariates. First, we calculate the principal components. The scree plot belows shows us that most of the variation in this data is captured by the top 7 principal components.

```{r}
# remove outlier sample
transformed_props <- props$TransformedProps[,colnames(props$TransformedProps) != "sample_23.1"]

PCs <- prcomp(t(transformed_props), center = TRUE, 
              scale = TRUE, retx = TRUE)
loadings = PCs$x # pc loadings
plot(PCs, type="lines") # scree plot
```

Collect all of the known sample traits.

```{r}
nGenes = nrow(transformed_props)
nSamples = ncol(transformed_props)

info <- seu@meta.data %>%
  dplyr::select(sample.id, 
                Disease,
                Treatment,
                Severity,
                Participant,
                Group_severity,
                Group,
                Batch, 
                Age, 
                Sex) %>%
  group_by(sample.id) %>%
  mutate(ncells = n()) %>%
  ungroup() %>%
  distinct()

m <- match(colnames(transformed_props), info$sample.id)
info <- info[m,]

datTraits <- info %>% dplyr::select(Participant, Batch, Disease, Treatment,
                                    Group, Severity, Age, Sex, ncells) %>%
  mutate(Age = log2(Age),
         ncells = log2(ncells),
    Donor = factor(Participant),
         Batch = factor(Batch),
         Disease = factor(Disease, 
                          labels = 1:length(unique(Disease))),
         Group = factor(Group, 
                        labels = 1:length(unique(Group))),
         Treatment = factor(Treatment, 
                            labels = 1:length(unique(Treatment))),
         Sex = factor(Sex, labels = length(unique(Sex))),
         Severity = factor(Severity, labels = length(unique(Severity)))) %>%
  mutate(across(everything(), as.numeric)) %>%
  dplyr::select(-Participant)

datTraits %>% 
  knitr::kable()
```

Correlate known sample traits with the top 10 principal components. This can help us determine which traits are potentially contributing to the main sources of variation in the data and should thus be included in our statistical analysis. 


```{r, fig.asp=1}
moduleTraitCor <- suppressWarnings(cor(loadings[, 1:10], datTraits, use = "p"))
moduleTraitPvalue <- WGCNA::corPvalueStudent(moduleTraitCor, (nSamples - 2))

textMatrix <- paste(signif(moduleTraitCor, 2), "\n(", 
                    signif(moduleTraitPvalue, 1), ")", sep = "")
dim(textMatrix) <- dim(moduleTraitCor)

## Display the correlation values within a heatmap plot
par(cex=0.75, mar = c(6, 8.5, 3, 3))
WGCNA::labeledHeatmap(Matrix = t(moduleTraitCor),
                      xLabels = colnames(loadings)[1:10],
                      yLabels = names(datTraits),
                      colorLabels = FALSE,
                      colors = WGCNA::blueWhiteRed(6),
                      textMatrix = t(textMatrix),
                      setStdMargins = FALSE,
                      cex.text = 1,
                      zlim = c(-1,1),
                      main = paste("PCA-trait relationships: Top 10 PCs"))
```

# Statistical analysis using `propeller` and `limma`

Create the design matrix.

```{r}
group <- factor(info$Group_severity)
participant <- factor(info$Participant)
age <- log2(info$Age)
batch <- factor(info$Batch)
sex <- factor(info$Sex)

design <- model.matrix(~ 0 + group + batch + age + sex)
colnames(design)[1:7] <- levels(group) 
design
```

Create the contrast matrix.

```{r}
contr <- makeContrasts(CF.NO_MODvNON_CF.CTRL = 0.5*(CF.NO_MOD.M + CF.NO_MOD.S) - NON_CF.CTRL,
                       CF.IVAvCF.NO_MOD = 0.5*(CF.IVA.S + CF.IVA.M) - 0.5*(CF.NO_MOD.S + CF.NO_MOD.M),
                       CF.LUMA_IVAvCF.NO_MOD = 0.5*(CF.LUMA_IVA.S + CF.LUMA_IVA.M) - 0.5*(CF.NO_MOD.S + CF.NO_MOD.M),
                       CF.NO_MOD.SvCF.NO_MOD.M = CF.NO_MOD.S - CF.NO_MOD.M,
                       levels = design)
contr
```

Add random effect for samples from the same individual.

```{r}
dupcor <- duplicateCorrelation(transformed_props, design=design,
                                block=participant)
dupcor
```

Fit the model.

```{r}
fit <- lmFit(transformed_props, design=design, block=participant, 
                correlation=dupcor$consensus)
fit2 <- contrasts.fit(fit, contr)
fit2 <- eBayes(fit2, robust=TRUE, trend=FALSE)
pvalue <- 0.05
summary(decideTests(fit2, p.value = pvalue))
```

## Results

```{r, fig.asp=1.25, fig.width=8, results='asis'}
p <- vector("list", ncol(contr))

for(i in 1:ncol(contr)){
  print(knitr::kable(topTable(fit2, coef = i),
                     caption = colnames(contr)[i]))
  
  transformed_props %>% data.frame %>%
    left_join(info,
              by = c("sample" = "sample.id")) %>%
    mutate(Group = Group_severity) %>%
    dplyr::filter(Group %in% names(contr[,i])[abs(contr[, i]) > 0]) -> dat
  
  if(length(unique(dat$Group)) > 2) dat$Group <- str_remove(dat$Group, ".(M|S)$")
  ggplot(dat, aes(x = Group,
                  y = Freq,
                  colour = Group,
                  group = Group)) +
    geom_jitter(stat = "identity",
                width = 0.15,
                size = 2) +
    stat_summary(geom = "point",
      fun.y = "mean",
      col = "black",
      shape = "_",
      size = 14) +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 90,
                                     hjust = 1,
                                     vjust = 0.5),
          legend.position = "bottom",
          legend.direction = "horizontal") +
    labs(x = "Group", y = "Proportion",
         colour = "Condition") +
    facet_wrap(~clusters, scales = "free_y") +
    ggtitle(colnames(contr)[i]) -> p[[i]]
  
  print(p[[i]])
}
```



# Session info {.appendix}

<summary>The analysis and this document were prepared using the following software (click triangle to expand)</summary>
<details>

```{r}
sessioninfo::session_info()
```

</details>










