---
title: "Inflammation of Paediatric Pulmonary Diseases"
subtitle: "Cell type proportions analysis: macrophages, annotation level 3"
author: "Jovana Maksimovic"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
site: workflowr::wflow_site
output:
  bookdown::html_document2:
    base_format: workflowr::wflow_html
    toc: yes
    toc_float: yes
    theme: cosmo
    highlight: textmate
    number_sections: true
---

```{r, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

# Load libraries

```{r, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(edgeR)
  library(tidyverse)
  library(ggplot2)
  library(Seurat)
  library(glmGamPoi)
  library(dittoSeq)
  library(clustree)
  library(AnnotationDbi)
  library(org.Hs.eg.db)
  library(glue)
  library(speckle)
  library(patchwork)
  library(paletteer)
  library(tidyHeatmap)
  library(here)
})

set.seed(42)
options(scipen=999)
options(future.globals.maxSize = 6500 * 1024^2)
```

# Load Data

```{r}
file <- here("data",
            "C133_Neeland_merged",
            glue("C133_Neeland_full_clean_macrophages_annotated_diet.SEU.rds"))

seu <- readRDS(file)
seu
```

```{r, echo=FALSE}
# cleanup obsolete objects
rm(seuLst)
gc()
```


# Analyse Cell type proportions 

```{r}
# Differences in cell type proportions
props <- getTransformedProps(clusters = seu$ann_level_3,
                             sample = seu$sample.id, transform="asin")
props$Proportions %>% knitr::kable()
```

## Cell type proportions by sample

Create sample meta data table.

```{r}
seu@meta.data %>%
  dplyr::select(sample.id,
                Participant,
                Disease,
                Treatment,
                Severity,
                Group,
                Group_severity,
                Batch, 
                Age, 
                Sex) %>%
     left_join(props$Counts %>% 
                 data.frame %>%
                 group_by(sample) %>%
                 summarise(ncells = sum(Freq)),
               by = c("sample.id" = "sample")) %>%
    distinct() -> info

head(info) %>% knitr::kable()
```


```{r, fig.asp=0.75, fig.width=9}
props$Proportions %>%
  data.frame %>%
  left_join(info, 
             by = c("sample" = "sample.id")) %>%
ggplot(aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", color = "black", size = 0.1) +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5,
                                   hjust = 1),
        legend.text = element_text(size = 8),
        legend.position = "bottom") +
  labs(y = "Proportion", fill = "Cell Label") +
  scale_fill_paletteer_d("Polychrome::glasbey", direction = -1) +
  facet_grid(~Group, scales = "free_x", space = "free_x")
```

## No. cells per sample

```{r, fig.asp=0.5, fig.width=9}
info %>%
ggplot(aes(x = sample.id, y = ncells, fill = Disease)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5,
                                   hjust = 1),
        legend.text = element_text(size = 8),
        legend.position = "bottom") +
  labs(y = "No. cells", fill = "Disease") +
  facet_grid(~Group, scales = "free_x", space = "free_x") +
  geom_hline(yintercept = 100, linetype = "dashed")
```

## Cell proportions by cell type

```{r}
props$Proportions %>%
  data.frame %>%
  left_join(info, 
             by = c("sample" = "sample.id")) %>%
ggplot(aes(x = clusters, y = Freq, fill = clusters)) +
  geom_boxplot(outlier.size = 0.1, size = 0.25) +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1),
        legend.text = element_text(size = 8)) +
  labs(y = "Proportion") +
  scale_fill_paletteer_d("Polychrome::glasbey", direction = -1) +
  NoLegend()
```


# Explore sources of variation
## Cell count data

Look at the sources of variation in the raw cell count level data.

```{r, fig.asp=1}
dims <- list(c(1,2), c(2:3), c(3,4), c(4,5))
p <- vector("list", length(dims))

for(i in 1:length(dims)){
  mds <- plotMDS(props$Counts,
    gene.selection = "common",
    plot = FALSE, dim.plot = dims[[i]])
  
  data.frame(x = mds$x, 
             y = mds$y,
             sample = rownames(mds$distance.matrix.squared)) %>%
    left_join(info, 
             by = c("sample" = "sample.id")) %>%
    distinct() -> dat
  
  p[[i]] <- ggplot(dat, aes(x = x, y = y, 
                            shape = as.factor(Disease),
                            color = as.factor(Batch))) +
    geom_point(size = 3) +
    labs(x = glue("Principal Component {dims[[i]][1]}"),
         y = glue("Principal Component {dims[[i]][2]}"),
         colour = "Batch",
         shape = "Disease") +
    theme(legend.direction = "horizontal",
          legend.text = element_text(size = 8),
          legend.title = element_text(size = 9),
          axis.text = element_text(size = 8),
          axis.title = element_text(size = 9)) 
}

wrap_plots(p, cols = 2) + plot_layout(guides = "collect") &
  theme(legend.position = "bottom") 
```

```{r, fig.asp=1}
dims <- list(c(1,2), c(2:3), c(3,4), c(4,5))
p <- vector("list", length(dims))

for(i in 1:length(dims)){
  mds <- plotMDS(props$Counts, 
                 gene.selection = "common",
                 plot = FALSE, dim.plot = dims[[i]])
  
  data.frame(x = mds$x, 
             y = mds$y,
             sample = rownames(mds$distance.matrix.squared)) %>%
    left_join(info, 
             by = c("sample" = "sample.id")) %>%
    distinct() -> dat
  
  p[[i]] <- ggplot(dat, aes(x = x, y = y, 
                            colour = log2(ncells)))+
    geom_text(aes(label = str_remove_all(sample, "sample_")), size = 2.5) +
    labs(x = glue("Principal Component {dims[[i]][1]}"),
         y = glue("Principal Component {dims[[i]][2]}"),
         colour = "Log2 No. Cells") +
    theme(legend.direction = "horizontal",
          legend.text = element_text(size = 8),
          legend.title = element_text(size = 9),
          axis.text = element_text(size = 8),
          axis.title = element_text(size = 9)) +
    scale_colour_viridis_c(option = "magma")
}

wrap_plots(p, cols = 2) + plot_layout(guides = "collect") &
  theme(legend.position = "bottom") 
```

## Cell proportion data

Look at the sources of variation in the cell proportions data.

```{r, fig.asp=1}
dims <- list(c(1,2), c(2:3), c(3,4), c(4,5))
p <- vector("list", length(dims))

for(i in 1:length(dims)){
  mds <- plotMDS(props$TransformedProps,
    gene.selection = "common",
    plot = FALSE, dim.plot = dims[[i]])
  
  data.frame(x = mds$x, 
             y = mds$y,
             sample = rownames(mds$distance.matrix.squared)) %>%
    left_join(info, 
             by = c("sample" = "sample.id")) %>%
    distinct() -> dat
  
  p[[i]] <- ggplot(dat, aes(x = x, y = y, 
                            shape = as.factor(Disease),
                            color = as.factor(Batch)))+
    geom_point(size = 3) +
    labs(x = glue("Principal Component {dims[[i]][1]}"),
         y = glue("Principal Component {dims[[i]][2]}"),
         colour = "Batch",
         shape = "Disease") +
    theme(legend.direction = "horizontal",
          legend.text = element_text(size = 8),
          legend.title = element_text(size = 9),
          axis.text = element_text(size = 8),
          axis.title = element_text(size = 9))
}

wrap_plots(p, cols = 2) + plot_layout(guides = "collect") &
  theme(legend.position = "bottom") 
```

```{r, fig.asp=1}
dims <- list(c(1,2), c(2:3), c(3,4), c(4,5))
p <- vector("list", length(dims))

for(i in 1:length(dims)){
  mds <- plotMDS(props$TransformedProps,
    gene.selection = "common",
    plot = FALSE, dim.plot = dims[[i]])
  
  data.frame(x = mds$x, 
             y = mds$y,
             sample = rownames(mds$distance.matrix.squared)) %>%
    left_join(info, 
             by = c("sample" = "sample.id")) %>%
    distinct() -> dat
  
  p[[i]] <- ggplot(dat, aes(x = x, y = y, 
                            shape = as.factor(Disease),
                            color = Sex))+
    geom_point(size = 3) +
    labs(x = glue("Principal Component {dims[[i]][1]}"),
         y = glue("Principal Component {dims[[i]][2]}"),
         colour = "Sex",
         shape = "Disease") +
    theme(legend.direction = "horizontal",
          legend.text = element_text(size = 8),
          legend.title = element_text(size = 9),
          axis.text = element_text(size = 8),
          axis.title = element_text(size = 9))
}

wrap_plots(p, cols = 2) + plot_layout(guides = "collect") &
  theme(legend.position = "bottom") 
```

```{r, fig.asp=1}
dims <- list(c(1,2), c(2:3), c(3,4), c(4,5))
p <- vector("list", length(dims))

for(i in 1:length(dims)){
  mds <- plotMDS(props$TransformedProps, 
                 gene.selection = "common",
                 plot = FALSE, dim.plot = dims[[i]])
  
   data.frame(x = mds$x, 
             y = mds$y,
             sample = rownames(mds$distance.matrix.squared)) %>%
    left_join(info, 
             by = c("sample" = "sample.id")) %>%
    distinct() -> dat
  
  p[[i]] <- ggplot(dat, aes(x = x, y = y, 
                            colour = log2(Age)))+
    geom_text(aes(label = str_remove_all(sample, "sample_")), size = 2.5) +
    labs(x = glue("Principal Component {dims[[i]][1]}"),
         y = glue("Principal Component {dims[[i]][2]}"),
         colour = "Log2 Age") +
    theme(legend.direction = "horizontal",
          legend.text = element_text(size = 8),
          legend.title = element_text(size = 9),
          axis.text = element_text(size = 8),
          axis.title = element_text(size = 9)) +
    scale_colour_viridis_c(option = "magma")
}

wrap_plots(p, cols = 2) + plot_layout(guides = "collect") &
  theme(legend.position = "bottom") 
```

```{r, fig.asp=1}
dims <- list(c(1,2), c(2:3), c(3,4), c(4,5))
p <- vector("list", length(dims))

for(i in 1:length(dims)){
  mds <- plotMDS(props$TransformedProps, 
                 gene.selection = "common",
                 plot = FALSE, dim.plot = dims[[i]])
  
  data.frame(x = mds$x, 
             y = mds$y,
             sample = rownames(mds$distance.matrix.squared)) %>%
    left_join(info, 
             by = c("sample" = "sample.id")) %>%
    distinct() -> dat
  
  p[[i]] <- ggplot(dat, aes(x = x, y = y, 
                            colour = log2(ncells)))+
    geom_text(aes(label = str_remove_all(sample, "sample_")), size = 2.5) +
    labs(x = glue("Principal Component {dims[[i]][1]}"),
         y = glue("Principal Component {dims[[i]][2]}"),
         colour = "Log2 No. Cells") +
    theme(legend.direction = "horizontal",
          legend.text = element_text(size = 8),
          legend.title = element_text(size = 9),
          axis.text = element_text(size = 8),
          axis.title = element_text(size = 9)) +
    scale_colour_viridis_c(option = "magma")
}

wrap_plots(p, cols = 2) + plot_layout(guides = "collect") &
  theme(legend.position = "bottom") 
```


# Principal components versus traits

Principal components analysis (PCA) allows us to mathematically determine the sources of variation in the data. We can then investigate whether these correlate with any of the specifed covariates. First, we calculate the principal components. The scree plot belows shows us that most of the variation in this data is captured by the top 7 principal components.

```{r}
# remove outlying sample
info <- info[info$sample.id != "sample_23.1",]
props$TransformedProps <- props$TransformedProps[, colnames(props$TransformedProps) != "sample_23.1"]

PCs <- prcomp(t(props$TransformedProps), center = TRUE, 
              scale = TRUE, retx = TRUE)
loadings = PCs$x # pc loadings
plot(PCs, type="lines") # scree plot
```

Collect all of the known sample traits.

```{r}
nGenes = nrow(props$TransformedProps)
nSamples = ncol(props$TransformedProps)

m <- match(colnames(props$TransformedProps), info$sample.id)
info <- info[m,]

datTraits <- info %>% dplyr::select(Participant, Batch, Disease, Treatment,
                                    Group, Severity, Age, Sex, ncells) %>%
  mutate(Age = log2(Age),
         ncells = log2(ncells),
    Donor = factor(Participant),
         Batch = factor(Batch),
         Disease = factor(Disease, 
                          labels = 1:length(unique(Disease))),
         Group = factor(Group, 
                        labels = 1:length(unique(Group))),
         Treatment = factor(Treatment, 
                            labels = 1:length(unique(Treatment))),
         Sex = factor(Sex, labels = length(unique(Sex))),
         Severity = factor(Severity, labels = length(unique(Severity)))) %>%
  mutate(across(everything(), as.numeric)) %>%
  dplyr::select(-Participant)

datTraits %>% 
  knitr::kable()
```

Correlate known sample traits with the top 10 principal components. This can help us determine which traits are potentially contributing to the main sources of variation in the data and should thus be included in our statistical analysis. 


```{r, fig.asp=1}
moduleTraitCor <- suppressWarnings(cor(loadings[, 1:10], datTraits, use = "p"))
moduleTraitPvalue <- WGCNA::corPvalueStudent(moduleTraitCor, (nSamples - 2))

textMatrix <- paste(signif(moduleTraitCor, 2), "\n(", 
                    signif(moduleTraitPvalue, 1), ")", sep = "")
dim(textMatrix) <- dim(moduleTraitCor)

## Display the correlation values within a heatmap plot
par(cex=0.75, mar = c(6, 8.5, 3, 3))
WGCNA::labeledHeatmap(Matrix = t(moduleTraitCor),
                      xLabels = colnames(loadings)[1:10],
                      yLabels = names(datTraits),
                      colorLabels = FALSE,
                      colors = WGCNA::blueWhiteRed(6),
                      textMatrix = t(textMatrix),
                      setStdMargins = FALSE,
                      cex.text = 1,
                      zlim = c(-1,1),
                      main = paste("PCA-trait relationships: Top 10 PCs"))
```

# Statistical analysis using `propeller` and `limma`

Create the design matrix.

```{r}
group <- factor(info$Group_severity)
participant <- factor(info$Participant)
age <- log2(info$Age)
batch <- factor(info$Batch)
sex <- factor(info$Sex)

design <- model.matrix(~ 0 + group + batch + age + sex)
colnames(design)[1:7] <- levels(group) 
design
```

Create the contrast matrix.

```{r}
contr <- makeContrasts(CF.NO_MODvNON_CF.CTRL = 0.5*(CF.NO_MOD.M + CF.NO_MOD.S) - NON_CF.CTRL,
                       CF.IVAvCF.NO_MOD = 0.5*(CF.IVA.S + CF.IVA.M) - 0.5*(CF.NO_MOD.S + CF.NO_MOD.M),
                       CF.LUMA_IVAvCF.NO_MOD = 0.5*(CF.LUMA_IVA.S + CF.LUMA_IVA.M) - 0.5*(CF.NO_MOD.S + CF.NO_MOD.M),
                       CF.NO_MOD.SvCF.NO_MOD.M = CF.NO_MOD.S - CF.NO_MOD.M,
                       levels = design)
contr
```

Add random effect for samples from the same individual.

```{r}
dupcor <- duplicateCorrelation(props$TransformedProps, design=design,
                                block=participant)
dupcor
```

Fit the model.

```{r}
fit <- lmFit(props$TransformedProps, design=design, block=participant, 
                correlation=dupcor$consensus)
fit2 <- contrasts.fit(fit, contr)
fit2 <- eBayes(fit2, robust=TRUE, trend=FALSE)
pvalue <- 0.05
summary(decideTests(fit2, p.value = pvalue))
```

## Results

```{r, fig.asp=1.5, fig.width=8, results='asis'}
p <- vector("list", ncol(contr))

for(i in 1:ncol(contr)){
  print(knitr::kable(topTable(fit2, coef = i, number = Inf),
                     caption = colnames(contr)[i]))
  
  props$Proportions %>% data.frame %>%
    left_join(info,
              by = c("sample" = "sample.id")) %>%
    mutate(Group = Group_severity) %>%
    dplyr::filter(Group %in% names(contr[,i])[abs(contr[, i]) > 0]) -> dat
  
  if(length(unique(dat$Group)) > 2) dat$Group <- str_remove(dat$Group, ".(M|S)$")
  ggplot(dat, aes(x = Group,
                  y = Freq,
                  colour = Group,
                  group = Group)) +
    geom_jitter(stat = "identity",
                width = 0.15,
                size = 2) +
    stat_summary(geom = "point",
      fun.y = "mean",
      col = "black",
      shape = "_",
      size = 14) +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 90,
                                     hjust = 1,
                                     vjust = 0.5),
          legend.position = "bottom",
          legend.direction = "horizontal") +
    labs(x = "Group", y = "Proportion",
         colour = "Condition") +
    facet_wrap(~clusters, scales = "free_y", ncol = 4) +
    ggtitle(colnames(contr)[i]) -> p[[i]]
  
  print(p[[i]])
}
```



# Session info {.appendix}

<summary>The analysis and this document were prepared using the following software (click triangle to expand)</summary>
<details>

```{r}
sessioninfo::session_info()
```

</details>










