---
title: "Figure 1"
author: "Jovana Maksimovic"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
site: workflowr::wflow_site
output:
  bookdown::html_document2:
    base_format: workflowr::wflow_html
    toc: yes
    toc_float: yes
    theme: cosmo
    highlight: textmate
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries.

```{r}
suppressPackageStartupMessages({
 library(SingleCellExperiment)
 library(edgeR)
 library(tidyverse)
 library(ggplot2)
 library(Seurat)
 library(glmGamPoi)
 library(dittoSeq)
 library(here)
 library(clustree)
 library(patchwork)
 library(AnnotationDbi)
 library(org.Hs.eg.db)
 library(glue)
 library(speckle)
 library(tidyHeatmap)
 library(paletteer)
 library(dsb)
 library(ggh4x)
 library(readxl)
})

source(here("code/utility.R"))
```

# Prepare figure panels
## ADT heatmap

Load the ADT data from save, pre-processed objects.

```{r}
files <- list.files(here("data/C133_Neeland_merged"),
                    pattern = "C133_Neeland_full_clean.*(macrophages|t_cells|other_cells)_annotated_full.SEU.rds",
                    full.names = TRUE)

# read in the ADT data, excluding everything else
seuLst <- lapply(files, function(f){
  s <- readRDS(f)
  DefaultAssay(s) <- "ADT"
  DietSeurat(s, assays = "ADT", dimreducs = NULL)
})

adt_names <- rownames(seuLst[[1]][["ADT"]]@counts)
seuLst <- lapply(seuLst, function(s){
  DefaultAssay(s) <- "ADT"
  
  if(!all(rownames(s) == adt_names)){
    adt_counts <- s[["ADT"]]@counts
    rownames(adt_counts) <- adt_names
    CreateSeuratObject(counts = adt_counts,
                       assay = "ADT",
                       meta.data = s@meta.data)
    
  } else {
    s
  }
})

seuADT <- merge(seuLst[[1]], 
                y = c(seuLst[[2]], 
                      seuLst[[3]]))
seuADT <- seuADT[, seuADT$Batch != 0]
seuADT
```

Normalise the ADT data across all cells using DSB.

```{r}
# read in TotalSeq feature information
read_csv(file = here("data",
                     "C133_Neeland_batch1",
                     "data",
                     "sample_sheets",
                     "ADT_features.csv")) -> adt_data
pattern <- "anti-human/mouse |anti-human/mouse/rat |anti-mouse/human "
adt_data$name <- gsub(pattern, "", adt_data$name)

out <- here("data",
            "C133_Neeland_merged",
            glue("C133_Neeland_full_clean_all_cells_dsb.ADT.rds"))
if(!file.exists(out)){
  adt_data %>%
    dplyr::filter(grepl("[Ii]sotype", name)) %>%
    pull(name) -> isotype_controls
  
  # normalise ADT using DSB normalisation
  adt_dsb <- ModelNegativeADTnorm(cell_protein_matrix = seuADT[["ADT"]]@counts,
                                  denoise.counts = TRUE,
                                  use.isotype.control = TRUE,
                                  isotype.control.name.vec = isotype_controls)
  saveRDS(adt_dsb, file = out)
  
} else {
  adt_dsb <- readRDS(out)
  
}

seuADT[["ADT"]]@data <- adt_dsb
seuADT
```
Set up mapping for long cell type labels to short labels.

```{r}
labels <- unique(seuADT$ann_level_3)

# Make a lookup table: label -> ann_level_1/2/3
hier_lut <- tibble(label = labels) %>%
  mutate(
    ann_level_1 = case_when(
      str_detect(label, "^macro") ~ "myeloid",
      label %in% c("monocytes","neutrophil-like","mast cells") ~ "myeloid",
      label %in% c("cDC1","cDC2","plasmacytoid DC","migratory DC") ~ "myeloid",
      label %in% c("CD4 T cells","CD8 T-rm","CD8 T-GZMK","CD8 T-inflammasome",
                   "CD4 T-rm","CD4 T-NFKB","CD4 T-naïve","CD4 T-IFN","CD4 T-reg",
                   "gamma delta T cells","NK-T cells","proliferating T/NK") ~ "lymphoid",
      label %in% c("NK cells","innate lymphocytes","dividing innate cells") ~ "lymphoid",
      label %in% c("B cells","HSP+ B cells","plasma B cells") ~ "lymphoid",
      label %in% c("secretory epithelial cells","ciliated epithelial cells") ~ "epithelial",
      TRUE ~ "other"
    ),
    ann_level_2 = case_when(
      # Myeloid
      str_detect(label, "^macro") ~ "macrophages",
      label %in% c("monocytes") ~ "monocytes",
      label %in% c("neutrophil-like", "mast cells") ~ "granulocytes",
      label %in% c("cDC1","cDC2","plasmacytoid DC","migratory DC") ~ "dendritic cells",

      # Lymphoid
      label %in% c("CD4 T cells","CD4 T-rm","CD4 T-NFKB","CD4 T-naïve","CD4 T-IFN","CD4 T-reg",
                   "CD8 T-rm","CD8 T-GZMK","CD8 T-inflammasome","gamma delta T cells","NK-T cells",
                   "proliferating T/NK") ~ "T cells",
      label %in% c("NK cells","innate lymphocytes","dividing innate cells") ~ "innate lymphoid",
      label %in% c("B cells","HSP+ B cells","plasma B cells") ~ "B cells",

      # Epithelial
      label %in% c("secretory epithelial cells","ciliated epithelial cells") ~ "epithelial cells",
      TRUE ~ "other"
    ),
    ann_level_3 = case_when(
      # Macrophage subtypes
      label == "macro-alveolar" ~ "macro-alveolar",
      str_detect(label, "^macro-alveolar") ~ label,  # keep specific alveolar flavors
      label %in% c("macro-interstitial","macro-monocyte-derived","macro-T") ~ label,
      label %in% c("macro-proliferating-S","macro-proliferating-G2M") ~ "macro-proliferating",

      # DC subtypes
      label %in% c("cDC1","cDC2","plasmacytoid DC","migratory DC") ~ label,

      # Neutro/mono/mast
      label %in% c("monocytes","neutrophil-like","mast cells") ~ label,

      # T cells
      label == "CD4 T cells" ~ "CD4 T cells",
      label %in% c("CD4 T-rm","CD4 T-NFKB","CD4 T-naïve","CD4 T-IFN","CD4 T-reg") ~ label,
      label %in% c("CD8 T-rm","CD8 T-GZMK","CD8 T-inflammasome") ~ label,
      label %in% c("gamma delta T cells","NK-T cells") ~ label,
      label == "proliferating T/NK" ~ "proliferating T-NK",

      # NK / ILC
      label %in% c("NK cells","innate lymphocytes","dividing innate cells") ~ label,

      # B lineage
      label %in% c("B cells","HSP+ B cells","plasma B cells") ~ label,

      # Epithelia
      label %in% c("secretory epithelial cells","ciliated epithelial cells") ~ label,

      TRUE ~ label
    )
  )

# Inspect:
hier_lut
```

Map long labels to short labels.

```{r}
# mapping to get short labels
lab_map <- c(
  "DC cells"                  = "DC",
  "mast cells"                = "Mast",
  "B cells"                   = "B",
  "monocytes"                 = "Mono",
  "epithelial cells"          = "Epi",
  "neutrophils"               = "Neut",
  "dividing innate cells"     = "Div innate",
  "NK cells"                  = "NK",
  "CD4 T cells"               = "CD4 T",
  "CD8 T cells"               = "CD8 T",
  "innate lymphocyte"         = "ILC",
  "gamma delta T cells"       = "γδ T",
  "NK-T cells"                = "NKT",
  "proliferating T/NK"        = "Prolif T/NK",
  "macrophages"               = "Mac",
  "proliferating macrophages" = "Prolif Mac"
)

# map long labels to short labels
seuADT$short_labels <- lab_map[seuADT$ann_level_1]
# match ordering of the levels betwen long and short labels
lut <- unique(seuADT@meta.data[, c("ann_level_1", "short_labels")])
lut <- lut[match(levels(factor(seuADT$ann_level_1)), lut$ann_level_1), , drop = FALSE]
# update level ordering for short labels
seuADT$short_labels <- factor(seuADT$short_labels, levels = lut$short_labels)
```


Make data frame of proteins, clusters, expression levels. Examine distribution of expression for heatmap scaling.

```{r}
# ADTs <- read_csv(file = here("data",
#                        "Proteins_broad_22.04.22.csv"))
# pattern <- "anti-human/mouse |anti-human/mouse/rat |anti-mouse/human |anti-human "
# ADTs$Description <- gsub(pattern, "", ADTs$Description)

labels <- readxl::read_excel(here("data/main_proteins.xlsx"))

unnest(enframe(setNames(str_split(labels$`main proteins`, ", "),
                        labels$`cell type`),
               value = "ADT",
               name = "cluster"),
       cols = ADT) %>%
  arrange(cluster) %>%
  distinct() -> markers

markers <- markers[markers$ADT %in% rownames(seuADT),]

seuADT@meta.data %>%
  dplyr::select(short_labels) %>%
  rownames_to_column(var = "cell") %>%
  inner_join(as.data.frame(t(seuADT[["ADT"]]@data)) %>%
               rownames_to_column(var = "cell")) %>%
  pivot_longer(c(-cell, -short_labels),
               names_to = "ADT",
               values_to = "Expression") %>%
  dplyr::group_by(short_labels, ADT) %>%
  dplyr::summarize(Expression = mean(Expression)) %>%
  ungroup() %>%
  dplyr::filter(ADT %in% markers$ADT) -> dat

plot(density(dat$Expression))
```


```{r}
dat %>%
  dplyr::rename("Protein" = "ADT",
                "ADT Exp." = "Expression",
                "Cell type" = "short_labels") %>%
  tidyHeatmap::heatmap(
    .column = Protein,
    .row = `Cell type`,
    .value = `ADT Exp.`,
    scale = "none",
    rect_gp = grid::gpar(col = "white", lwd = 1),
    show_row_names = TRUE, 
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    column_names_gp = grid::gpar(fontsize = 8, fontfamily = "arial"),
    column_title_gp = grid::gpar(fontsize = 10, fontfamily = "arial"),
    row_names_gp = grid::gpar(fontsize = 8, fontfamily = "arial"),
    row_title_gp = grid::gpar(fontsize = 10, fontfamily = "arial"),
    column_title_side = "top",
    palette_value = circlize::colorRamp2(seq(0, 2, length.out = 11),
                                            rev(RColorBrewer::brewer.pal(11, "RdYlBu"))),
    heatmap_legend_param = list(direction = "vertical")) %>%
  add_tile(`Cell type`, show_legend = FALSE,
           show_annotation_name = FALSE,
           palette = paletteer_d("miscpalettes::pastel", 
                                 length(unique(seuADT$ann_level_1)))) %>%
    as_ComplexHeatmap() -> f1e
f1e
```


## UMAP of all cells

Split by *batch* for integration. Normalise with `SCTransform`. Increase the strength of alignment by increasing `k.anchor` parameter to 20 as recommended in Seurat Fast integration with RPCA [vignette](https://satijalab.org/seurat/articles/integration_rpca.html).

```{r}
# clean up memory
gc()

out <- here("data",
            "C133_Neeland_merged",
            glue("C133_Neeland_full_clean_integrated_all_cells.SEU.rds"))

if(!file.exists(out)){
  
  # load annotated cells for each cell type group
  files <- list.files(here("data/C133_Neeland_merged"),
                      pattern = "C133_Neeland_full_clean.*(macrophages|t_cells|other_cells)_annotated_diet.SEU.rds",
                      full.names = TRUE)
  seuLst <- lapply(files[2:4], function(f) readRDS(f))
  # merge into a single object
  seu <- merge(seuLst[[1]],
               y = c(seuLst[[2]],
                     seuLst[[3]]))
  rm(seuLst)
  gc()
  
  # Assign each cell a score, based on its expression of G2/M and S phase markers as described in the Seurat workflow
  # https://satijalab.org/seurat/articles/cell_cycle_vignette.html
  s.genes <- cc.genes.updated.2019$s.genes
  g2m.genes <- cc.genes.updated.2019$g2m.genes
  seu <- CellCycleScoring(seu, s.features = s.genes, g2m.features = g2m.genes, 
                          set.ident = TRUE)

  # Using the `Seurat` *Alternate Workflow* from (https://satijalab.org/seurat/articles/cell_cycle_vignette.html),
  # calculate the difference between the G2M and S phase scores so that signals separating non-cycling cells and cycling
  # cells will be maintained, but differences in cell cycle phase among proliferating cells (which are often
  # uninteresting), can be regressed out of the data.
  seu$CC.Difference <- seu$S.Score - seu$G2M.Score
  
  gns <- AnnotationDbi::select(org.Hs.eg.db,
                               keys = rownames(seu),
                               columns = c("CHR","ENTREZID"),
                               keytype = "SYMBOL",
                               multiVals = "first")
  
  m <- match(rownames(seu), gns$SYMBOL)
  gns <- gns[m,]
  # remove HLA, immunoglobulin, MT, RP, MRP and sex genes prior to integration
  var_regex = '^HLA-|^IG[HJKL]|^MT-|^RPL|^MRPL'
  keep <- !(str_detect(rownames(seu), var_regex) | gns$CHR %in% c("X","Y"))
  seu <- seu[keep,] 

  DefaultAssay(seu) <- "RNA"
  VariableFeatures(seu) <- NULL
  
  seuLst <- SplitObject(seu, split.by = "Batch")
  rm(seu)
  gc()
  
  # normalise with SCTransform and regress out cell cycle score difference
  seuLst <- lapply(X = seuLst, FUN = SCTransform, method = "glmGamPoi",
                   vars.to.regress = "CC.Difference")
  # integrate RNA data
  features <- SelectIntegrationFeatures(object.list = seuLst,
                                        nfeatures = 3000)
  seuLst <- PrepSCTIntegration(object.list = seuLst, anchor.features = features)
  seuLst <- lapply(X = seuLst, FUN = RunPCA, features = features)
  anchors <- FindIntegrationAnchors(object.list = seuLst,
                                    normalization.method = "SCT",
                                    anchor.features = features,
                                    dims = 1:30, reduction = "rpca")
  seu <- IntegrateData(anchorset = anchors, 
                       normalization.method = "SCT",
                       dims = 1:30)
  
  DefaultAssay(seu) <- "integrated"
  seu <- RunPCA(seu, dims = 1:30, verbose = FALSE) %>%
    RunUMAP(dims = 1:30, verbose = FALSE)

  saveRDS(seu, file = out)
  fs::file_chmod(out, "664")
  if(any(str_detect(fs::group_ids()$group_name, 
                    "oshlack_lab"))) fs::file_chown(out, 
                                                    group_id = "oshlack_lab")
} else {
  seu <- readRDS(file = out)
  
}
```

Map long cell type labels to short labels for RNA data.

```{r}
# map long labels to short labels
seu$short_labels <- lab_map[seu$ann_level_1]
# match ordering of the levels betwen long and short labels
lut <- unique(seu@meta.data[, c("ann_level_1", "short_labels")])
lut <- lut[match(levels(factor(seu$ann_level_1)), lut$ann_level_1), , drop = FALSE]
# update level ordering for short labels
seu$short_labels <- factor(seu$short_labels, levels = lut$short_labels)
```


```{r, fig.asp=1}
options(ggrepel.max.overlaps = Inf)
cluster_pal <- "miscpalettes::pastel"

DimPlot(seu, group.by = "Group")
FeaturePlot(seu, features = "Age") +
  scale_color_viridis_c(option = "magma", direction = -1)

DimPlot(seu, 
        group.by = "short_labels", label = F) +
  scale_color_paletteer_d(cluster_pal) +
  NoLegend() -> p1

LabelClusters(p1, id = "short_labels", repel = TRUE,
              size = 2.5, box = TRUE,
              fontface = "bold") +
  theme(axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        plot.title = element_blank()) -> f1b

f1b
```

## Cell proportions by sample

```{r, fig.asp=0.75, fig.width=10}
samp_map <-
c(
  "CF.IVA" = "CF (iva)",
  "CF.LUMA_IVA" = "CF (luma/iva)",
  "CF.NO_MOD" = "CF (no mod)",
  "NON_CF.CTRL" = "Non-CF control"
)

seu@meta.data %>%
  dplyr::select(sample.id, Group) %>%
  mutate(Group = samp_map[Group]) %>%
  count(sample.id, Group) %>% 
  ungroup() %>%
ggplot(aes(x = sample.id, y = n, fill = Group)) +
  geom_bar(stat = "identity", color = "black", size = 0.1) +
  theme_classic() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.line.x = element_blank(),
        strip.text = element_blank(),
        strip.background = element_blank(),
        plot.margin = unit(c(0, 0, 0, 0), "lines")) +
  labs(y = "Number of cells", fill = "Condition") +
  scale_fill_paletteer_d("RColorBrewer::Set2", direction = 1) +
  facet_grid(~Group, scales = "free_x", space = "free_x") +
  scale_y_continuous(expand = expansion(mult = c(0.01, 0.02))) -> p2

props <- getTransformedProps(clusters = seu$short_labels,
                             sample = seu$sample.id, transform="asin")

props$Proportions %>%
  data.frame %>%
  inner_join(seu@meta.data %>%
               dplyr::select(sample.id,
                             Group),
             by = c("sample" = "sample.id")) %>%
  distinct() %>%
ggplot(aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", color = "black", size = 0.1) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size = 8),
        strip.text = element_blank(),
        strip.background = element_blank(),
        plot.margin = unit(c(0, 0, 0, 0), "lines")) +
  labs(y = "Cell type proportion", fill = "Cell type", x = "Sample") +
  scale_fill_paletteer_d("miscpalettes::pastel", direction = 1) +
  facet_grid(~Group, scales = "free_x", space = "free_x") +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.03))) -> p3

(p2 / p3) + plot_layout(guides = "collect") &
  theme(legend.text = element_text(size = 8),
        legend.title = element_text(size = 10),
        legend.key.size = unit(0.8, "lines")) -> f1c

f1c
```

## `Seurat` marker gene dotplot

```{r}
DefaultAssay(seu) <- "RNA"
Idents(seu) <- "ann_level_1"

gns <- AnnotationDbi::select(org.Hs.eg.db,
                             keys = rownames(seu),
                             columns = c("CHR","ENTREZID"),
                             keytype = "SYMBOL",
                             multiVals = "first")
m <- match(rownames(seu), gns$SYMBOL)
gns <- gns[m,]

out <- here("data/cluster_annotations/seurat_markers_all_cells.rds")

if(!file.exists(out)){
  keep <- !is.na(gns$ENTREZID)
  markers <- FindAllMarkers(seu, only.pos = TRUE, logfc.threshold = 0.5,
                            features = rownames(seu)[rownames(seu) %in% gns$SYMBOL[keep]],
                            max.cells.per.ident = 10000)
  saveRDS(markers, file = out)

} else {
  markers <- readRDS(out)

}

# labels <- readxl::read_excel(here("data/main_marker_genes.xlsx"))
# 
# unnest(enframe(setNames(str_split(labels$`main marker genes`, ", "),
#                         labels$`cell type`),
#                value = "gene",
#                name = "cluster"),
#        cols = gene) %>%
#   arrange(cluster) %>%
#   distinct() -> markers

markers <- markers[markers$gene %in% rownames(seu),]
```

```{r, fig.width=12, fig.asp=0.4}
draw_marker_gene_dotplot(seu,
                         markers %>% mutate(cluster = as.character(cluster)),
                         ann_level = "ann_level_1",
                         cluster_pal,
                         lab_map = lab_map,
                         direction = 1,
                         num = 5) -> f1d

f1d
```

## Cell type dendrogram

```{r}
library(dplyr)
library(data.tree)
library(igraph)
library(ggraph)

# df has ann_level_1/2/3 (one row per cell)
paths <- (seu@meta.data %>% 
            dplyr::select(ann_level_3) %>%
            dplyr::rename(label = ann_level_3)) %>%
  left_join(hier_lut) %>%
  mutate(across(starts_with("ann_level_"), ~ tidyr::replace_na(.x, "unassigned"))) %>%
  count(ann_level_1, ann_level_2, ann_level_3, name = "cells") %>%
  mutate(pathString = paste("all cells", ann_level_1, ann_level_2, ann_level_3, sep = "/"))


# ----- 2) tree & propagate counts -----
tree <- as.Node(paths, pathName = "pathString")
tree$Do(function(node) {
  if (!node$isLeaf) node$cells <- sum(sapply(node$children, function(ch) ifelse(is.null(ch$cells), 0, ch$cells)))
}, traversal = "post-order")

# ----- 3) nodes/edges + lineage & clean labels -----
nodes <- ToDataFrameTree(tree, "levelName", "pathString", "cells", "level") %>%
  transmute(
    name       = pathString,  # unique vertex id (full path)
    short_name = str_squish(str_replace(levelName, "^[^A-Za-z]+", "")),  # strip only leading non-letters
    cells      = cells,
    depth      = level,
    lineage    = sub("^all cells/([^/]+).*", "\\1", name)  # ann_level_1 from path segment 2
  )

edges <- nodes %>%
  mutate(parent = ifelse(name == "all cells", NA_character_, dirname(name))) %>%
  filter(!is.na(parent)) %>%
  dplyr::select(from = parent, to = name)

g <- graph_from_data_frame(edges, vertices = nodes, directed = TRUE)

# mark leaves once (optional; ggraph usually provides `leaf` automatically for dendrograms)
V(g)$leaf <- igraph::degree(g, mode = "out") == 0

ggraph(g, layout = "dendrogram") +
  geom_edge_diagonal(color = "lightgrey") +
  geom_node_point(aes(size = cells, colour = lineage)) +
  geom_node_text(aes(label = short_name),
                 angle = 0, 
                 hjust = 0, 
                 size = 3, 
                 nudge_y = 0.05,
                 show.legend = FALSE) +
  scale_size_continuous(name = "No. cells", range = c(1, 6)) +
  # Root on left, leaves on right, with padding on BOTH sides
  coord_flip(clip = "off") +  # KEY: no clipping
  scale_y_reverse(expand = expansion(mult = c(0.02, 0.3))) +
  theme_void() +
  theme(plot.margin = margin(rep(0, 4))) +
  labs(colour = "Lineage") -> f1f

f1f
```

# Figure 1

```{r, fig.asp=1, fig.width=13}
layout = "
BBBCCCCC
BBBCCCCC
BBBCCCCC
BBBCCCCC
DDDDDDDD
DDDDDDDD
DDDDDDDD
DDDDDDDD
FFFFGGGG
FFFFGGGG
FFFFGGGG
FFFFGGGG
FFFFGGGG
"
(wrap_elements(f1b + theme(plot.margin = unit(rep(0,4), "cm"))) + 
    wrap_elements(f1c + theme(plot.margin = unit(rep(0,4), "cm"))) + 
    wrap_elements(f1d + theme(plot.margin = unit(rep(0,4), "cm"))) + 
    wrap_plots(list(f1e %>% 
                      ComplexHeatmap::draw(heatmap_legend_side = "left") %>% 
                      grid::grid.grabExpr())) +
    wrap_elements(f1f + theme(plot.margin = unit(rep(0,4), "cm")))) + 
  plot_layout(design = layout) +
  plot_annotation(tag_levels = list(c("B","C","D","E","F"))) &
  theme(plot.tag = element_text(size = 24,
                                face = "bold",
                                family = "arial"))
  
```



# Session info




