---
title: "Figure 1"
author: "Jovana Maksimovic"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
site: workflowr::wflow_site
output:
  bookdown::html_document2:
    base_format: workflowr::wflow_html
    toc: yes
    toc_float: yes
    theme: cosmo
    highlight: textmate
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries.

```{r}
suppressPackageStartupMessages({
 library(SingleCellExperiment)
 library(edgeR)
 library(tidyverse)
 library(ggplot2)
 library(Seurat)
 library(glmGamPoi)
 library(dittoSeq)
 library(here)
 library(clustree)
 library(patchwork)
 library(AnnotationDbi)
 library(org.Hs.eg.db)
 library(glue)
 library(speckle)
 library(tidyHeatmap)
 library(paletteer)
 library(dsb)
 library(ggh4x)
  library(readxl)
})

source(here("code/utility.R"))
```

# Load data

```{r}
files <- list.files(here("data/C133_Neeland_merged"),
                    pattern = "C133_Neeland_full_clean.*(macrophages|t_cells|other_cells)_annotated_diet.SEU.rds",
                    full.names = TRUE)

seuLst <- lapply(files[2:4], function(f) readRDS(f))

seu <- merge(seuLst[[1]], 
             y = c(seuLst[[2]], 
                   seuLst[[3]]))
seu
```

```{r}
rm(seuLst)
gc()
```

# Prepare figure panels
## Data integration

Visualise batch effects.

```{r, message=FALSE, warning=FALSE, fig.asp=0.9}
seu <- ScaleData(seu) %>%
  FindVariableFeatures() %>%
  RunPCA(dims = 1:30, verbose = FALSE) %>%
  RunUMAP(dims = 1:30, verbose = FALSE)
```

```{r}
DimPlot(seu, group.by = "Batch", reduction = "umap")

#cluster_pal <- "ggsci::category20_d3"
cluster_pal <- "miscpalettes::pastel"
DimPlot(seu, group.by = "ann_level_1", reduction = "umap") +
  theme(legend.direction = "vertical",
        legend.text = element_text(size = 10)) +
  scale_color_paletteer_d(palette = cluster_pal)
```


## Cell cycle effect

Assign each cell a score, based on its expression of G2/M and S phase markers as described in the Seurat workflow [here](https://satijalab.org/seurat/articles/cell_cycle_vignette.html).

```{r, warning=FALSE}
s.genes <- cc.genes.updated.2019$s.genes
g2m.genes <- cc.genes.updated.2019$g2m.genes

seu <- CellCycleScoring(seu, s.features = s.genes, g2m.features = g2m.genes, 
                        set.ident = TRUE)
```

PCA of cell cycle genes.

```{r, fig.asp=1.25}
DimPlot(seu, group.by = "Phase") -> p1

seu %>%
  RunPCA(features = c(s.genes, g2m.genes),
                      dims = 1:30, verbose = FALSE) %>%
  DimPlot(reduction = "pca") -> p2

(p2 / p1) + plot_layout(guides = "collect")
```

Distribution of cell cycle markers.

```{r, fig.asp=0.8}
# Visualize the distribution of cell cycle markers across
RidgePlot(seu, features = c("PCNA", "TOP2A", "MCM6", "MKI67"), ncol = 2,
          log = TRUE)
```

Using the `Seurat` *Alternate Workflow* from [here](https://satijalab.org/seurat/articles/cell_cycle_vignette.html), calculate the difference between the G2M and S phase scores so that signals separating non-cycling cells and cycling cells will be maintained, but differences in cell cycle phase among proliferating cells (which are often uninteresting), can be regressed out of the data.

```{r}
seu$CC.Difference <- seu$S.Score - seu$G2M.Score
```


## Integrate RNA data

Split by *batch* for integration. Normalise with `SCTransform`. Increase the strength of alignment by increasing `k.anchor` parameter to 20 as recommended in Seurat Fast integration with RPCA [vignette](https://satijalab.org/seurat/articles/integration_rpca.html).

First, integrate the RNA data.

```{r}
out <- here("data",
            "C133_Neeland_merged",
            glue("C133_Neeland_full_clean_integrated_all_cells.SEU.rds"))

gns <- AnnotationDbi::select(org.Hs.eg.db,
                             keys = rownames(seu),
                             columns = c("CHR","ENTREZID"),
                             keytype = "SYMBOL",
                             multiVals = "first")
m <- match(rownames(seu), gns$SYMBOL)
gns <- gns[m,]
# remove HLA, immunoglobulin, MT, RP, MRP and sex genes prior to integration
var_regex = '^HLA-|^IG[HJKL]|^MT-|^RPL|^MRPL'
keep <- !(str_detect(rownames(seu), var_regex) | gns$CHR %in% c("X","Y"))
seu <- seu[keep,] 
  
if(!file.exists(out)){
  DefaultAssay(seu) <- "RNA"
  VariableFeatures(seu) <- NULL
  seu[["pca"]] <- NULL
  seu[["umap"]] <- NULL
  
  seuLst <- SplitObject(seu, split.by = "Batch")
  rm(seu)
  gc()
  
  # normalise with SCTransform and regress out cell cycle score difference
  seuLst <- lapply(X = seuLst, FUN = SCTransform, method = "glmGamPoi",
                   vars.to.regress = "CC.Difference")
  # integrate RNA data
  features <- SelectIntegrationFeatures(object.list = seuLst,
                                        nfeatures = 3000)
  seuLst <- PrepSCTIntegration(object.list = seuLst, anchor.features = features)
  seuLst <- lapply(X = seuLst, FUN = RunPCA, features = features)
  anchors <- FindIntegrationAnchors(object.list = seuLst,
                                    normalization.method = "SCT",
                                    anchor.features = features,
                                    dims = 1:30, reduction = "rpca")
  seu <- IntegrateData(anchorset = anchors, 
                       normalization.method = "SCT",
                       dims = 1:30)
  
  DefaultAssay(seu) <- "integrated"
  seu <- RunPCA(seu, dims = 1:30, verbose = FALSE) %>%
    RunUMAP(dims = 1:30, verbose = FALSE)

  saveRDS(seu, file = out)
  fs::file_chmod(out, "664")
  if(any(str_detect(fs::group_ids()$group_name, 
                    "oshlack_lab"))) fs::file_chown(out, 
                                                    group_id = "oshlack_lab")
  
} else {
  seu <- readRDS(file = out)
  
}
```

```{r}
seu <- subset(seu, cells = which(seu$ann_level_2 != "macro-T"))
```

```{r, fig.asp=1}
options(ggrepel.max.overlaps = Inf)
DimPlot(seu, 
        group.by = "ann_level_1", label = F, repel = T,
        label.size = 3) +
  scale_color_paletteer_d(cluster_pal, direction = 1) +
  NoLegend() -> p1

LabelClusters(p1, id = "ann_level_1", repel = TRUE, 
              size = 2, box = TRUE, fontfamily = "arial") +
  theme(axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        plot.title = element_blank()) -> f1b

f1b
```


## Cell proportions by sample

```{r, fig.asp=0.75, fig.width=10}
seu@meta.data %>%
  dplyr::select(sample.id, Group) %>%
  count(sample.id, Group) %>% 
  ungroup() %>%
ggplot(aes(x = sample.id, y = n, fill = Group)) +
  geom_bar(stat = "identity", color = "black", size = 0.1) +
  theme_classic() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.line.x = element_blank(),
        strip.text = element_blank(),
        strip.background = element_blank(),
        plot.margin = unit(c(0, 0, 0, 0), "lines")) +
  labs(y = "Number of cells", fill = "Condition") +
  scale_fill_paletteer_d("RColorBrewer::Set2", direction = 1) +
  facet_grid(~Group, scales = "free_x", space = "free_x") -> p2

props <- getTransformedProps(clusters = seu$ann_level_1,
                             sample = seu$sample.id, transform="asin")

props$Proportions %>%
  data.frame %>%
  inner_join(seu@meta.data %>%
               dplyr::select(sample.id,
                             Group),
             by = c("sample" = "sample.id")) %>%
  distinct() %>%
ggplot(aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", color = "black", size = 0.1) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size = 8),
        strip.text = element_blank(),
        strip.background = element_blank(),
        plot.margin = unit(c(0, 0, 0, 0), "lines")) +
  labs(y = "Cell type proportion", fill = "Cell type", x = "Sample") +
  scale_fill_paletteer_d("miscpalettes::pastel", direction = 1) +
  facet_grid(~Group, scales = "free_x", space = "free_x") -> p3

(p2 / p3) + plot_layout(guides = "collect") &
  theme(legend.text = element_text(size = 8),
        legend.title = element_text(size = 10),
        legend.key.size = unit(1, "lines")) -> f1c

f1c
```

```{r}
DefaultAssay(seu) <- "RNA"
Idents(seu) <- "ann_level_1"

gns <- AnnotationDbi::select(org.Hs.eg.db,
                             keys = rownames(seu),
                             columns = c("CHR","ENTREZID"),
                             keytype = "SYMBOL",
                             multiVals = "first")
m <- match(rownames(seu), gns$SYMBOL)
gns <- gns[m,]

out <- here("data/cluster_annotations/seurat_markers_all_cells.rds")

if(!file.exists(out)){
  keep <- !is.na(gns$ENTREZID)
  markers <- FindAllMarkers(seu, only.pos = TRUE, logfc.threshold = 0.5,
                            features = rownames(seu)[rownames(seu) %in% gns$SYMBOL[keep]],
                            max.cells.per.ident = 10000)
  saveRDS(markers, file = out)

} else {
  markers <- readRDS(out)

}

# labels <- readxl::read_excel(here("data/main_marker_genes.xlsx"))
# 
# unnest(enframe(setNames(str_split(labels$`main marker genes`, ", "),
#                         labels$`cell type`),
#                value = "gene",
#                name = "cluster"),
#        cols = gene) %>%
#   arrange(cluster) %>%
#   distinct() -> markers

markers <- markers[markers$gene %in% rownames(seu),]
```

## `Seurat` marker gene dotplot

```{r, fig.width=12, fig.asp=0.4}
draw_marker_gene_dotplot(seu,
                         markers,
                         "ann_level_1",
                         cluster_pal,
                         direction = 1,
                         num = 5) -> f1d

f1d
```

```{r}
files <- list.files(here("data/C133_Neeland_merged"),
                    pattern = "C133_Neeland_full_clean.*(macrophages|t_cells|other_cells)_annotated_full.SEU.rds",
                    full.names = TRUE)

seuLst <- lapply(files[2:4], function(f) readRDS(f))

adt_names <- rownames(seuLst[[1]][["ADT"]]@counts)
seuLst <- lapply(seuLst, function(s){
  DefaultAssay(s) <- "ADT"
  
  if(!all(rownames(s) == adt_names)){
    adt_counts <- s[["ADT"]]@counts
    rownames(adt_counts) <- adt_names
    CreateSeuratObject(counts = adt_counts,
                       assay = "ADT",
                       meta.data = s@meta.data)
    
  } else {
    DietSeurat(s, assays = "ADT", dimreducs = NULL)
    
  }
  
})

seuADT <- merge(seuLst[[1]], 
                y = c(seuLst[[2]], 
                      seuLst[[3]]))
seuADT <- seuADT[, seuADT$Batch != 0]
seuADT
```

## Visualise ADTs
Make data frame of proteins, clusters, expression levels.

```{r}
out <- here("data",
            "C133_Neeland_merged",
            glue("C133_Neeland_full_clean_all_cells_dsb.ADT.rds"))

read_csv(file = here("data",
                     "C133_Neeland_batch1",
                     "data",
                     "sample_sheets",
                     "ADT_features.csv")) -> adt_data
pattern <- "anti-human/mouse |anti-human/mouse/rat |anti-mouse/human "
adt_data$name <- gsub(pattern, "", adt_data$name)

if(!file.exists(out)){
  adt_data %>%
    dplyr::filter(grepl("[Ii]sotype", name)) %>%
    pull(name) -> isotype_controls
  
  # normalise ADT using DSB normalisation
  adt_dsb <- ModelNegativeADTnorm(cell_protein_matrix = seuADT[["ADT"]]@counts,
                                  denoise.counts = TRUE,
                                  use.isotype.control = TRUE,
                                  isotype.control.name.vec = isotype_controls)
  saveRDS(adt_dsb, file = out)
  
} else {
  adt_dsb <- readRDS(out)
  
}

seuADT[["ADT"]]@data <- adt_dsb
seuADT
```

```{r}
# ADTs <- read_csv(file = here("data",
#                        "Proteins_broad_22.04.22.csv"))
# pattern <- "anti-human/mouse |anti-human/mouse/rat |anti-mouse/human |anti-human "
# ADTs$Description <- gsub(pattern, "", ADTs$Description)

labels <- readxl::read_excel(here("data/main_proteins.xlsx"))

unnest(enframe(setNames(str_split(labels$`main proteins`, ", "),
                        labels$`cell type`),
               value = "ADT",
               name = "cluster"),
       cols = ADT) %>%
  arrange(cluster) %>%
  distinct() -> markers

markers <- markers[markers$ADT %in% rownames(seuADT),]

seuADT@meta.data %>%
  dplyr::select(ann_level_1) %>%
  rownames_to_column(var = "cell") %>%
  inner_join(as.data.frame(t(seuADT[["ADT"]]@data)) %>%
               rownames_to_column(var = "cell")) %>%
  pivot_longer(c(-cell, -ann_level_1),
               names_to = "ADT",
               values_to = "Expression") %>%
  dplyr::group_by(ann_level_1, ADT) %>%
  dplyr::summarize(Expression = mean(Expression)) %>%
  ungroup() %>%
  dplyr::filter(ADT %in% markers$ADT) -> dat

plot(density(dat$Expression))
```

```{r}
dat %>%
  dplyr::rename("Protein" = "ADT",
                "ADT Exp." = "Expression",
                "Cell type" = "ann_level_1") %>%
  tidyHeatmap::heatmap(
    .column = Protein,
    .row = `Cell type`,
    .value = `ADT Exp.`,
    scale = "none",
    rect_gp = grid::gpar(col = "white", lwd = 1),
    show_row_names = TRUE, 
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    column_names_gp = grid::gpar(fontsize = 8, fontfamily = "arial"),
    column_title_gp = grid::gpar(fontsize = 10, fontfamily = "arial"),
    row_names_gp = grid::gpar(fontsize = 8, fontfamily = "arial"),
    row_title_gp = grid::gpar(fontsize = 10, fontfamily = "arial"),
    column_title_side = "top",
    palette_value = circlize::colorRamp2(seq(0, 2, length.out = 11),
                                         rev(RColorBrewer::brewer.pal(11, "Spectral"))),
    heatmap_legend_param = list(direction = "vertical")) %>%
  add_tile(`Cell type`, show_legend = FALSE,
           show_annotation_name = FALSE,
           palette = paletteer_d("miscpalettes::pastel", 
                                 length(unique(seuADT$ann_level_1)))) %>%
    as_ComplexHeatmap() -> f1e
f1e
```


# Figure 1

```{r, fig.asp=1, fig.width=13}
layout = "
BBBCCCCC
BBBCCCCC
BBBCCCCC
DDDDDDDD
DDDDDDDD
FFFFGGGG
FFFFGGGG
"
(wrap_elements(f1b + theme(plot.margin = unit(rep(0,4), "cm"))) + 
    wrap_elements(f1c + theme(plot.margin = unit(rep(0,4), "cm"))) + 
    wrap_elements(f1d + theme(plot.margin = unit(rep(0,4), "cm"))) + 
    wrap_plots(list(f1e %>% 
                      ComplexHeatmap::draw(heatmap_legend_side = "right") %>% 
                      grid::grid.grabExpr())) +
    plot_spacer()) + 
  plot_layout(design = layout) +
  plot_annotation(tag_levels = list(c("B","C","D","E"))) &
  theme(plot.tag = element_text(size = 16,
                                face = "bold",
                                family = "arial"))
  
```

# Session info





<!-- ```{r, fig.width=10, fig.asp=0.8} -->
<!-- seu@meta.data %>% -->
<!--     data.frame() %>% -->
<!--     ggplot(aes(x = sample.id, fill = Group)) + -->
<!--     geom_bar(position = position_dodge()) +  -->
<!--     geom_text(aes(label = ..count..), -->
<!--               stat = "count", vjust = -0.5,  -->
<!--               colour = "black", size = 2) + -->
<!--   facet_wrap(~Batch, scales = "free_x", ncol = 4) + -->
<!--   theme_classic() + -->
<!--   theme(axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5)) + -->
<!--   labs(x = "Sample") -->
<!-- ``` -->


<!-- ```{r} -->
<!-- seu@meta.data %>% -->
<!--     data.frame() %>% -->
<!--   dplyr::select(sample.id, Group) %>% -->
<!--   distinct() %>% -->
<!--     ggplot(aes(x = Group, fill = Group)) + -->
<!--     geom_bar(position = position_dodge()) +  -->
<!--     geom_text(aes(label = ..count..), -->
<!--               stat = "count", vjust = -0.5,  -->
<!--               colour = "black", size = 3) + -->
<!--   theme_classic() + -->
<!--   theme(legend.position = "none") + -->
<!--   labs(x = "Condition", y = "No. Samples") -->
<!-- ``` -->



