---
title: "Getting Started"
author: "Jovana Maksimovic"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
site: workflowr::wflow_site
output:
  bookdown::html_document2:
    base_format: workflowr::wflow_html
    toc: yes
    toc_float: yes
    theme: cosmo
    highlight: textmate
    number_sections: true
---

```{r knitr, include = FALSE}
knitr::opts_chunk$set(autodep        = TRUE,
                      cache          = FALSE,
                      cache.comments = FALSE,
                      echo           = TRUE,
                      error          = FALSE,
                      fig.align      = "center",
                      fig.width      = 10,
                      fig.height     = 8,
                      message        = FALSE,
                      warning        = FALSE)
```

This page describes how to download the data and code used in this analysis,
set up the project directory and reproduce the analysis. We have used the
[`workflowr`][workflowr] and [`renv`][renv] packages to organise this project and 
ensure reproducibility.

# Getting the code {#code}

All the code and outputs of this analysis are available from GitHub at
[https://github.com/Oshlack/paediatric-cf-inflammation-citeseq](https://github.com/Oshlack/paediatric-cf-inflammation-citeseq). If you want to replicate the analysis you can either clone the repository or 
download it as a zipped directory.

Once you have a local copy of the repository you should see the following
directory structure:

* `analysis/` - Contains the RMarkdown documents with the various stages of
analysis. These are numbered according to the order they should be run.
* `code/` - R scripts with custom functions used in some analysis stages. 
* `data/` - This directory contains the data files used in the analysis with
sub-directories for different data types (see [Getting the data](#data) for
details). Processed intermediate data files will also be placed here.
* `docs/` - This directory contains the analysis website html files hosted at
  [https://oshlacklab.com/paediatric-cf-inflammation-citeseq/](https://oshlacklab.com/paediatric-cf-inflammation-citeseq/), 
  as well as the image files.
* `output/` - Directory for output files produced by the analysis.
* `renv/` 
* `README.md` - README describing the project.
* `.Rprofile` - Custom R profile for the project including set up for
  `workflowr`.
* `.gitattributes`
* `.gitignore` - Details of files and directories that are excluded from the
  repository.
* `_workflowr.yml` - `workflowr` configuration file.
* `paediatric-cf-inflammation-citeseq.Rproj` - RStudio project file.
* `renv.lock` - `renv` lock file, used to restore and install correct versions of
R packages required for this project.

This analysis was completed using `r R.version.string`. To ensure reproducibility
the `renv` package was used to track package sources and versions. Ensure you 
have the correct version of R and `renv` installed prior to beginning. To 
install the necessary package versions you can use:

```r
renv::restore()
```

For more information on using `renv` see the `renv` [website][renv].


# Getting the data {#data}

**Note on data availability:** The data provided represents the earliest stage of processing that can be shared openly. Upstream steps (01.0–01.6) required genetic information from participants for sample demultiplexing; interested parties requiring access to the raw FASTQ files may contact the authors, and access may be arranged subject to appropriate institutional ethics and governance approvals. The preprocessed count data provided here has been generated after demultiplexing and does not contain genetic information.

The preprocessed single-cell RNA-seq and CITE-seq count data generated for this study can be downloaded from [Zenodo](https://doi.org/10.5281/zenodo.19364848). To unpack the data into the correct directory structure, place the archive in the root of the cloned repository and run:

`tar -xzf C133_Neeland_preprocessed_SCEs.tar.gz -C data`

on the command line.


The cell and gene quality filtered, raw and normalised RNA data can also be explored and downloaded from [CELLxGENE](https://cellxgene.cziscience.com/collections/30704a65-1b50-409b-9cf6-1f26bf644c99). 


# Running the analysis {#analysis}

The `analysis` directory contains the following analysis files:

```{r, echo=FALSE}
list.files(path = here::here("analysis"), pattern = "[01][0-9]_*")
```

As indicated by the numbering they should be run in this order. The analysis can be reproduced starting from the data in the [Getting the data](#data) section by running individual stages of the analysis starting from step 02.0, either by
providing the names of the file you want to run to `workflowr::wflow_build()` or
by manually knitting the document (for example using the 'Knit' button in
RStudio). Note, most parts of the analysis require outputs generated by a 
previous step and so will not run unless the preceding steps have already been 
executed.

[geo]: <https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE127465> "geo"
[workflowr]: <https://workflowr.github.io/workflowr/> "workflowr"
[renv]: <https://rstudio.github.io/renv/articles/renv.html> "renv"